A Titin Missense Variant Causes Atrial Fibrillation

Version 1

Abstract

Abstract

Rare and common genetic variants contribute to the risk of atrial fibrillation (AF). Although ion channels were among the first AF candidate genes identified, rare loss-of-function variants in structural genes such as TTN have also been implicated in AF pathogenesis partly by the development of an atrial myopathy, but the underlying mechanisms are poorly understood. While TTN truncating variants (TTNtvs) have been causally linked to arrhythmia and cardiomyopathy syndromes, the role of missense variants (mvs) remains unclear. We report that rare TTNmvs are associated with adverse clinical outcomes in AF patients and we have identified a mechanism by which a TTNmv (T32756I) causes AF. Modeling the TTN-T32756I variant using human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs) revealed that the mutant cells display aberrant contractility, increased activity of a cardiac potassium channel (KCNQ1, Kv7.1), and dysregulated calcium homeostasis without compromising the sarcomeric integrity of the atrial cardiomyocytes. We also show that a titin-binding protein, the Four-and-a-Half Lim domains 2 (FHL2), has increased binding with KCNQ1 and its modulatory subunit KCNE1 in the TTN-T32756I-iPSC-aCMs, enhancing the slow delayed rectifier potassium current (Iks). Suppression of FHL2 in mutant iPSC-aCMs normalized the Iks, supporting FHL2 as an Iks modulator. Our findings demonstrate that a single amino acid change in titin not only affects function but also causes ion channel remodeling and AF. These findings emphasize the need for high-throughput screening to evaluate the pathogenicity of TTNmvs and establish a mechanistic link between titin, potassium ion channels, and sarcomeric proteins that may represent a novel therapeutic target.

Introduction

Atrial fibrillation (AF), the most prevalent cardiac arrhythmia, affects more than 60 million people worldwide and is associated with increased risk for stroke, heart failure, and dementia, justifying it as a major public healthcare burden2,3. AF is characterized by irregular and often abnormally fast heart rates4,5. Over the last decade, tremendous progress has been made in understanding the genetic architecture of AF6–8. Genome-wide association studies have identified over 140 common loci associated with AF, while family-based studies have implicated rare variants primarily encoding ion channels9,10. Although AF has been traditionally classified as a ‘channelopathy’, variants in myocardial sarcomeric proteins such as titin have been increasingly associated with familial or early-onset AF11,12.

The TTN gene encodes for a massive myofilament (∼4200 kDa), titin, which stretches along the Z-disk (N-terminus) to the M-band (C-terminus) region of the sarcomere13,14. Titin serves as a molecular scaffold for other muscle proteins, participates in downstream signaling, and provides passive tension to cardiac muscle15,16. Due to its large size, the rate of genetic variation in TTN is high, including both truncating (TTNtvs) and missense variants (TTNmvs)1,17. The prevalence of rare TTNtvs and TTNmvs in the general population is 2% and 5.7%, respectively18–20. While TTNtvs are the most common genetic cause (10-20%) of dilated cardiomyopathy (DCM), TTNmvs are often disregarded in clinical practice, fulfilling a stand-alone benign criterion in DCM-specific variant interpretation frameworks based on the American College of Medical Genetics (ACMG) recommendations21,22. Yet, TTNmvs may have the potential to confer disease, as a recent study in two families revealed that TTNmvs in a conserved cysteine of TTN can cause DCM23. Another study demonstrated segregation of a novel TTNmv among five individuals with atrioventricular block in a Chinese family, suggesting that TTNmvs may also be implicated in arrhythmia syndromes24. To date, a relationship between rare TTNmvs and AF has not been explored at either clinical or mechanistic levels. This has potentially heightened importance across racial and ethnic groups, in whom the likelihood of detecting variants of uncertain significance (which are predominantly TTNmvs), is higher than in individuals of European descent25.

Despite the clinical importance, the pharmacological therapy of AF is limited in part because of the incomplete understanding of the myocardial substrate for AF as human atrial tissue is rarely available and the limitations of existing in vitro and in vivo models26. However, human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-CMs) not only possess the complex array of cardiac ion channels that make up the atrial action potential (AP) but also hold great promise for modeling AF27,28. Modeling patient-specific mutations associated with familial AF using mature human iPSC-CMs offers a powerful, naturally integrated system with distinct advantages over animal models and heterologous expression systems29. Recent reports using human iPSC-CM models have elucidated the pathophysiological mechanisms of DCM30, hypertrophic cardiomyopathy31, Brugada syndrome32, long QT syndrome33, and AF27,29,34. Human iPSC-CMs also have the potential to uncover the molecular mechanisms of AF as they can be easily modified to contain the precise genetic background of the individual patient35–37.

To examine the potential role of TTNmvs in AF, we examined the prevalence of rare TTNmvs in a single-center cohort of African-American and Hispanic/Latinx individuals with AF who underwent whole exome sequencing, compared clinical characteristics between TTNmv carriers and non-carriers, and evaluated whether TTNmvs were associated with increased risk of AF or heart failure (HF)-related hospitalizations. To explore whether the TTNmvs may be mechanistically linked to the development of AF, we introduced a rare TTNmv (TTN-T32756I) into human iPSC-aCMs using clustered regularly interspaced short palindromic repeats-associated 9 (CRISPR-Cas9). We performed electrophysiological (EP), pharmacological, and mechanistic analyses to elucidate the mechanisms by which the TTNmv causes AF. Our study showed that the TTN-T32756I iPSC-aCMs exhibited a striking AF-like EP phenotype in vitro, and transcriptomic analyses revealed that the TTNmv increases the activity of the Four-and-a-half LIM domain protein 2 (FHL2) which then modulates the slow delayed rectifier potassium current (IKs) to cause AF, possibly through a reentrant mechanism.

Results

TTNmvs are associated with increased hospitalization risk in a multiethnic cohort with AF

Clinical characteristics

A total of 131 subjects were included (mean [SD] age at AF diagnosis 63.5 [13.8] years, 70 [53.4%] male, 38 [29%] Hispanic/Latinx, 93 [71%] non-Hispanic Black; Table 1). We identified 138 TTNmvs, most commonly in the TTN A-band (108 [78.8]%), followed by the I-band (20 [14.6%]), M-band (2 [1.5%]), near the Z-disk (6 [4.4%], and in the Z-disk (2 [1.5%]) (Figure 1A, Supplementary Table 1). Based on the REVEL in silico score, 52 (37.7%) TTNmvs variants were predicted to be deleterious. A total of 77 (58.8%) subjects carried a TTNmv, with 43 (32.8%) carrying a predicted deleterious variant. Carriers of a TTNmv had a higher QTc interval on electrocardiogram (ECG) closest to AF diagnosis (mean [SD] 466.5 [42.3] vs. 449.6 [37.7] ms, P=0.027); there were otherwise no other significant clinical or demographic differences between variant carriers and non-carriers. When stratifying by predicted deleterious TTNmv only (Supplementary Table 2), a higher proportion of variant carriers were non-Hispanic Black (36 [83.7%] vs. 57 [64.8%], P=0.026). On index ECG, carriers additionally had higher ventricular rate (mean [SD] 103.7 [31.4] vs. 90.9 [27.4] beats/min, P=0.022), QTc interval (mean [SD] 470.6 [44.0] vs. 453.9 [38.7] ms, P=0.035), and LVEDD on index echocardiogram (mean [SD] 49.8 [8.0] vs. 45.6 [9.2] mm, P=0.021). Seventeen (15.2%) subjects met criteria for left ventricular (LV) dilatation, with a higher proportion of LV dilatation in TTNmv carriers but without a significant difference compared to non-carriers (13 [20.3%] vs. 4 [8.3%], P=0.111). Twelve subjects (9.2%) met criteria for a clinical diagnosis of nonischemic DCM based on a left ventricular ejection fraction (LVEF) <50%, presence of LV dilatation, and confirmation of non-severe coronary artery disease by coronary angiography. Of those, 8 subjects carried TTNmvs of which 6 were predicted to be deleterious (Supplementary Table 3).

TTNmv prevalence and association with hospitalization in a multiethnic atrial fibrillation (AF) cohort.

(A) Distribution of TTNmv in multiethnic AF cohort based on amino acid position in the TTN gene, stratified by REVEL in silico score for prediction of deleterious effect, defined by REVEL ≥ 0.70. (B) Mean cumulative incidence of AF and heart failure (HF)-related hospitalizations in subjects with AF stratified by presence of TTNmv. Hazard ratio (HR), 95% confidence interval (CI), and P-value were obtained from univariable Cox proportional hazard modeling. (C) Diagram denoting the location of TTNmv-T23756I. (D) Sequence alignment shows that the region of the T23756I is highly conserved across vertebrate species. (E) Allele frequencies of TTN-T3265I in various ethnic groups (gnomAD).

TTNmvs are associated with higher hospitalization risk

A total of 174 hospitalizations (64 AF-related and 110 HF-related) occurred after a median (interquartile range [IQR]) follow-up time of 4.14 (1.25-6.04) years, with 119 (68.4%) events occurring in TTNmv carriers and 55 (31.6%) of events occurring in non-TTNmv carriers. Thirty-nine (50.6%) subjects with a TTNmv experienced at least one hospitalization during the follow-up period compared to 22 (40.1%) of non-carriers, and total hospitalization incidence rate was 32.2 events per 100 person-years (p-y) in TTNmv carriers compared to 17.4 per 100 p-y in non-carriers. Mean cumulative incidence is shown in Figure 1B. The unadjusted hazard of hospitalization by 10 years was significantly higher in TTNmv carriers compared to noncarriers (hazard ratio [HR] 1.81, 95% confidence interval [CI] 1.04-3.15, p=0.036). This remained significant after partial adjustment for age and sex (HR 1.82, 95% CI 1.04-3.17, p=0.035) as well as full adjustment that additionally included ethnicity and baseline LVEF <50% (HR 1.80, 95% CI 1.03-3.15, p=0.039; Supplementary Table 4). No other covariates were significantly associated with the outcome. To assess whether baseline LVEF influenced the relationship between TTNmvs and hospitalization risk, an interaction term between LVEF<50% and presence of TTNmvs was tested and was not significant (interaction P=0.843). To assess whether prediction of deleterious effect influenced the relationship between TTNmv presence and hospitalization risk, subgroups based on high or low REVEL score were separately examined (Supplementary Table 5). Compared to non-TTNmv carriers, hospitalization risk was significantly increased in subjects carrying predicted deleterious TTNmvs (HR 1.92, 1.04-3.53, P=0.036). While subjects with TTNmvs not predicted to be deleterious also had an elevated point estimate of hospitalization risk, this difference was not significant (HR 1.60, 95% CI 0.78-3.28, P=0.198). Findings remained similar with exclusion of patients with nonischemic DCM (Supplementary Table 6). While in a previous study we identified likely pathogenic or pathogenic AF variants in 7.0% of ethnic minority probands with most (46.7%) TTNtvs12, these results suggest that TTNmvs are also linked with adverse clinical outcomes and missense variants should be considered in assessing pathogenicity in clinical and genetic contexts.

A single amino acid change in a TTNmv is potentially causative of AF

We identified three early-onset paroxysmal AF probands from separate families harboring the same rare TTNmv (Chr 2q31.2 (GRCh38): g. 178539797G>A). The clinical characteristics of the three probands are summarized in Table 2. This missense variant localizes in the A-band of TTN and causes the substitution of a conserved residue threonine by an isoleucine (p.Thr32756Ile, NM_001267550.2 (TTN), c.98267C>T (Figure 1C-D, Supplementary Figure 1A). In the gnomAD reference population, the TTN-T32756I variant has a frequency of 0.000245 overall, at a frequency of 0.0046 within the subpopulation of African ancestry and is impartially distributed across sexes and different ages (Figure 1E, Supplementary Figure 1B-E). The TTNmv has also been described in Clinvar (NM_001267550.2(TTN):c.98267C>T (p.Thr32756Ile), Supplementary Table 7) and gnomAD v4.0.0 (SNV:2-178539798-G-A(GRCh38)) in patients with DCM and other diseases.

TTN-T32756I iPSC-aCMs display aberrant contractility without affecting the sarcomere integrity

To functionally characterize the TTN-T32756I variant, we exploited wild-type (WT) human iPSC lines and introduced the point mutation c.98267C>T in the TTN allele (Figure 2A, Supplementary Figure 2A-B). The iPSC lines expressing the pluripotent markers Sox2 and Oct4 and showing normal karyotype were first differentiated into cardiomyocytes (Supplementary Figure 2C-D), and then a retinoic acid-based and comprehensive maturation protocol was applied to generate the iPSC-aCMs as we described previously26,27. Given that cardiac contractility is a key component of heart function and is linked to cardiac disorders associated with TTN38, we first assessed the contractility of the WT and mutant iPSC-aCMs and observed both reduced contraction and abnormal relaxation in TTN-T32756I-iPSC-aCMs (Figure 2B-E, Supplementary Figure 3A-B). Compared to WT, the beating frequency of the TTN-T32756I-iPSC-aCMs was significantly increased (52 ± 7.8 vs. 98 ± 7.5 beats per min, P=0.001; Figure 2C) coupled with the reduction of the contraction duration (456.5 ± 61.45 vs 262.9 ± 48.16 msec, P=0.032; Figure 2D), the peak-to-peak time (1529 ± 195.5 vs 636.6 ± 135.8 msec, P=0.004; Supplementary Figure 3B), and the relaxation (281.5 ± 42.95 vs 79.40 ± 21.14 msec, P=0.003; Supplementary Figure 3A). The contraction amplitude of the mutant was also increased (8.1 ± 0.8 vs 10.6 ± 7 au, P=0.040) without any significant changes in time-to-peak (Figure 2E, Supplementary Figure 3C), suggesting an increased contractile force by the TTN-T32756I-iPSC-aCMs. As sarcomere disorganization is often found in TTN-related cardiomyopathies and underlies contractile dysfunction39, we explored whether TTNmvs perturbed the organization of the sarcomere. Surprisingly, we did not find any disarray in the sarcomeres of either WT or TTN-32756I-iPSC-aCMs. Using transmission electron microscopy (TEM) and immunofluorescence (IF), we observed symmetrical arrays of recurring sarcomeres in the WT and the mutant atrial cardiomyocytes with no changes in the sarcomere length (Figure 2F-G, Supplementary Figure 3D-F). Although no sarcomeric perturbation by the T32756I variant was observed, the aberrant contractility displayed by the TTN-T32756I-iPSC-aCMs suggests that the TTNmv may be pathogenic.

Human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs) carrying TTN-T32756I variants have atypical contractility but no sarcomere disorganization.

(A) Workflow to generate the CRISPR/Cas9-mediated iPSC line carrying TTN-T32756I missense variation. (B-E) Contraction profile of wild type (black) and TTN-T32756I (Red) iPSC-aCMs showing increased beating frequency (C), Peak-to-Peak time (D), and Contraction duration (E) in the mutant. (F) Representative sarcomeric organization of wild-type (WT) and TTN-T32756I iPSC-aCM by Transmission electron microscopy (TEM). (G) There is no significant change in the sarcomere length (H). n.s.; P>0.05; *P<0.05; **P< 0.01.

Altered cardiac potassium current and calcium handling underlies the arrhythmia phenotype in TTN-T32756I-iPSC-aCMs

As studies have shown that abnormal atrial electrophysiology and calcium-handling underlie atrial arrhythmogenesis and contractile dysfunction40,41, we hypothesized that the TTN-T32756I variant modulates atrial APs, ion channels and intracellular calcium. We first tested the AP characteristics of the isolated WT and the TTN-T32756I iPSC-aCMs, which revealed a significant shortening of the AP duration (APD) at 10%, 50%, and 90% repolarization, however, there was no change in the peak amplitude (Figure 3A-C, Supplementary Figure 4A-B). Since increased potassium current (Ik) especially the augmented delayed rectifier potassium current (Iks) causes AF by reducing the APD27,42, we then assessed the Iks by whole-cell voltage clamping. We observed an increase in both Ik and Iks in TTN-T32756I-iPSC-aCMs when compared to the WT (Figure 3D-F, Supplementary Figure 4C-E); thus, confirming that the APD shortening is partly due to the increased Ik. To test the effect of T32756I on atrial calcium handling, we measured the intracellular calcium transients at the excitation-contraction coupling moment of the iPSC-aCMs using fluorescent (fura-2, AM) calcium imaging. Compared to the WT, TTN-T32756I-iPSC-aCMs exhibited increased arrhythmic frequency along with a significant reduction of the time to 50% and 90% decline of calcium transients (Figure 3G-I, Supplementary Figure 4F). However, TTN-T32756I-iPSC-CMs exhibited similar calcium transient amplitudes as the WT, indicating that there was no change in the availability of the intracellular calcium for each contraction (Supplementary Figure 4G). Our findings suggest that the reduced APD due to the increased potassium current and the decreased timing of the calcium transient may create a reentrant substrate for AF.

Effect of T32756I on action potential (AP) and calcium-handling in iPSC-aCMs.

(A-C) Representative optical AP recordings of WT and TTN-T32756I showing reduction of AP duration (APD) at the 50% (APD50) (B) and 90% (APD90) repolarization (C). (D) Current-voltage (I-V) curves of the slow delayed rectifier potassium current (Iks) in WT and TTN_T32756I iPSC-aCMs. control (n=7) (E-F) Comparison of Iks current density at 50 mV (mean ± SEM). N.s.; P>0.05; *P<0.05; **P<0.01; ***P<0.001; ****P<0.0001. (G) Representative tracings of spontaneous calcium transients of WT and TTN-T32756I iPSC-aCMs. (H-I) Calcium kinetics show that the TTN-T32756I iPSC-aCMs have increased frequency (B) and decreased transient durations (I) compared with the WT iPSC-aCMs.

RNA sequencing analysis reveals alterations in cardiac signaling and disease pathways in TTN-T32756I-iPSC-aCMs

To elucidate the underlying molecular mechanism by which TTN-T32756I causes contractile defects and ion channel remodeling, we performed transcriptome sequencing of the mutant iPSC-aCMs and WT. Comparison of total RNA levels in TTN-T32756-iPSC-aCMs with WT-iPSC-aCMs and differential expression analysis showed genes related to cardiac muscle contraction and calcium handling were predominantly affected (Figure 4A, Supplementary Figure 5A). Gene-ontology (GO) pathway enrichment further showed that the top significantly downregulated cardiac-related GO Biological Processes (-BP) in TTN-T32756I iPSC-aCMs included key processes such as cardiac myofibril assembly, skeletal muscle myosin thick filament assembly, extracellular matrix organization, and regulation of potassium ion transmembrane transporter assembly among others (Figure 4B). As for GO-Molecular function pathways (GO-MF), outward potassium channel activity, voltage-gated potassium channel activity, calcium channel activity, and gap junction channel activity were enriched (Figure 4C). Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway analysis identified downregulation in critical pathways including Adrenergic signaling in cardiomyocytes, the Phosphoinositide 3-kinase (PI3K) signaling pathway, Dilated cardiomyopathy pathway, Hypertrophic cardiomyopathy pathway and Calcium signaling pathway (Figure 4D), which are crucial for normal cardiac electrophysiology and contractility. Upregulated pathways in both GO and KEGG analyses were mostly restricted to neuronal pathways including axon and neuron development, Hippo signaling pathway, and Notch signaling pathway. In the transcriptomic analysis of TTN-T32756I-iPSC-aCMs, several transcription factors (TFs) showed significant alterations, highlighting their potential roles in the mutation’s impact (Figure 4E, Supplementary Figure 5D). Top-cardiac related upregulated TFs include MYC, linked to cell growth and apoptosis, suggesting an increase in cellular proliferation. Stress-related TFs such as NFE2L2 and CEBPB are also elevated, indicating a heightened stress response. Additionally, TFs like ESRRA and FOXM1, which regulate metabolic processes and cell cycle respectively, along with FOXO1 for metabolism and stress resistance, were upregulated. Key TFs in muscle function such as MYOD1, MYOCD, KLF5, and NKX2.5 were enhanced, alongside SMAD2, CREBBP, PITX2, NFKB1, and GATA4, which are vital for diverse roles ranging from signal transduction to immune regulation (Figure 4E). Conversely, downregulated transcription factors include SIRT1, which plays a role in cellular stress resistance and mitochondrial function, pointing to potential vulnerabilities in cellular defenses. Cardiovascular-related TFs such as KLF2 and KLF3, along with SMAD7, involved in TGF-beta signaling, were reduced, possibly impacting structural and signaling integrity. Additionally, MXD1 and IKZF1, crucial for cellular differentiation and immune development, were also diminished (Figure 4E). These transcriptional changes underscore a complex network of regulatory adjustments that could contribute to the altered electrophysiological and structural properties in TTN-T32756I-iPSC-aCMs.

Transcriptomic profile and pathway enrichment analysis comparing TTN-T32756I iPSC-aCMs with the WT.

(A) Heatmaps of cardiac-related upregulated and downregulated differentially expressed genes (DEGs) (B) Top significantly enriched downregulated cardiac-related Gene-Ontology Biological process (GO-BP) pathways in the TTN-T32756I iPSC-aCMs. (C) Top significantly enriched downregulated cardiac-related Gene-Ontology Molecular Function (GO-MF) pathways in the TTN-T32756I iPSC-aCMs. (D) Top significantly enriched downregulated cardiac-related Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways in the TTN-T32756I iPSC-aCMs.(E) Significantly enriched upregulated and downregulated transcription factors (TFs) (F) Network diagram showing the upregulation of KCNQ1 by FHL2 predicted by the Ingenuity pathway enrichment analysis (IPA).

Utilizing the KEGG and GO pathways, we performed Ingenuity Pathway Analysis (IPA) to identify novel titin-interacting proteins that could cause the increased Iks activity in the TTN-T32756I-iPSC-aCMs. The IPA suggests FHL2, which is known to transduce mechanical signaling through its interactions with titin and is enriched in the mutant iPSC-aCMs (Figure 4F). As FHL2 has also been shown to directly interact with the Iks-binding partner mink (KCNE1) and increases its current density43, we postulate that elevated FHL2 level in TTN-T32756I-iPSC-aCMs may cause enhanced binding to Iks, and thus increase its activity.

The enhanced Iks current in TTN-T32756I-iPSC-aCMs is mediated by the FHL2

To determine if TTN-T32756I increases Iks by modulating the interaction between KCNQ1-KCNE1 and FHL2, we performed co-immunoprecipitation studies in both WT and TTN-T32756I-iPSC-aCMs. The co-localization between KCNE1 and FHL2 increased ∼3 fold in TTN-T32756I-iPSC-aCMs, suggesting an increased interaction between them (Figure 5A). Since FHL2 enhances Iks activity, we then investigated whether inhibition of the FHL2 could reverse the increased Iks that was observed in the TTN-T32756I-iPSC-aCMs. To inhibit FHL2, we employed the small interfering (si) RNA to suppress the FHL2 gene and measured the Iks activity in both WT and TTN-T32756I-iPSC-aCMs. Both the WT and mutant iPSC-aCMs transfected with FHL2 specific siRNA showed a substantial decrease in FHL2 expression (Figure 5B). As shown in Figure 5C-D, voltage-clamp recordings demonstrate that the Iks in FHL2-suppressed TTN-T32756I-iPSC-aCMs is significantly reduced compared to corresponding the Iks in TTN-T32756I-iPSC-aCMs but comparable to the WT (Figure 5D). Therefore, inhibition of FHL2 by the siRNA rescues the increased Iks of the TTN-T32756I-iPSC-aCMs, which is similar to that of WT Iks. Overall, our data suggest that the TTNmv creates an EP substrate for AF by modulating the IKs activity in part by an increased interaction between the KCNQ1-KCNE1 complex and FHL2 (Figure 5E).

Inhibition of FHL2 rescues enhanced Iks in TTN-T32756I iPSC-aCMs.

(A) Co-immunoprecipitation revealed increased interaction between FHL2 and KCNQ1-KCNE1 (MinK) complex. (IP: KCNE1). Immunoblotting (IB) was performed with antibodies against FHL2 (32 kDa) and MinK (32 kDa) (n = 3) (B) qPCR data showing the inhibition of FHL2 gene by the siRNA in the WT and TTN-T32756I iPSC-aCMs (n=7). (C) I-V curves showing the rescue of the Iks TTN-T32756I iPSC-aCMs by the suppression of FHL2 (n=4-8). (D) Comparison of Iks current density at 50 mV (mean ± SEM). I Schematic showing the TTN-T32756I results in increased FHL2 binding with the KNCQ1-KCNE1 complex and enhanced Iks activity. *P<0.05, **P<0.01, ***P<0.001.

Discussion

We identified an association between TTNmvs and clinical outcomes among a multiethnic cohort of AF patients and elucidated a causal mechanism by which a TTNmv may lead to AF. Although TTNtvs are a well-recognized cause of DCM and have been associated with early-onset AF11,44, the role of TTNmvs in either arrhythmia or cardiomyopathy pathogenesis is unknown. Here, for the first time, we show that TTNmvs are associated with increased hospitalization risk in a multiethnic AF cohort, and that a TTNmv can cause AF by impairing contractility and ion channel remodeling in atrial cardiomyocytes. Furthermore, we demonstrate that augmented Iks mediated by FHL2 can create a substrate for AF. Together, these findings establish a potential causal role of TTNmvs in AF and support Iks and FHL2 as potential therapeutic targets for TTN-related AF.

The genetic architecture of AF is complex, with both monogenic and polygenic contributions that intersect with age and comorbidities to define an individual’s risk. Despite AF being traditionally considered an ion channelopathy, the sarcomeric gene, TTN, is the most common in which rare loss-of-function variants have been associated with AF11,45, and penetrance of such variants was found to be even greater for AF than for heart failure in the UK Biobank46. As rare TTN variants comprise the strongest monogenic contribution to AF risk, we sought to explore whether TTNmvs may similarly have implications in AF development or outcomes. A study of 147 probands with DCM identified 44 severe TTNmvs in 37 probands, which clustered in the A-band region1. There were no differences in heart transplant-free survival between carriers and noncarriers. In a more recent study of 530 subjects with DCM using more stringent allele frequency criteria, 31 predicted deleterious variants were identified, also predominantly located in the A-band19. However, these TTNmvs were found with similar frequency to reference populations. Finally, a study of two families with DCM demonstrated segregation of TTNmvs affecting the same highly conserved cysteine residue in the I-band, which showed impaired contraction and folding at physiological temperatures in a homozygous iPSC model, demonstrating that a TTNmv can cause DCM23.

We identified TTNmvs in 58% of subjects in our cohort, and 33% carried a potentially deleterious variant by REVEL in silico prediction. This frequency was higher than the 37/147 (25.2%) of TTNmv carriers in a DCM cohort17, understanding that differing methods of bioinformatic filtering may limit ability for direct comparison. Aligning with prior studies, most deleterious variants in our study were found in the A-band47,48, where TTNtvs have been linked most strongly to disease. Presence of predicted deleterious TTNmvs was associated with higher LV end diastolic diameter but without statistically significant differences in LVEF, potentially suggesting overlap between a subclinical left ventricular cardiomyopathy and AF in these subjects. We additionally noted higher ventricular rate and QTc interval on index ECGs in these subjects. Rare deleterious variants in TTN have previously been associated with changes in the QT interval49, but the mechanism for this remains unclear. Finally, evidence on the relationship between TTNmvs and clinical outcomes is sparse and limited to examining event-free survival in DCM patients. We observed a significant association between TTNmv presence and increased cumulative incidence of AF or HF-related hospitalizations, which remained unchanged in multivariable adjustment and in sensitivity analyses. This importantly suggests that TTNmvs may correlate with disease severity. Further study exploring these associations in larger validation cohorts and examining AF-specific measures such as arrhythmia burden or treatment response will be necessary to fully understand the clinical importance of TTNmvs in AF.

Prediction of missense variant impact using in silico tools and algorithms can be limited by low specificity, contributing to challenges in clinical interpretation50. Functional studies are often helpful in these cases to confirm or support variant pathogenicity. As existing in vitro or in vivo models do not adequately replicate the complexity of AF, human iPSC-aCMs possess the complex array of cardiac ion channels that make up the atrial AP and provide an adequate model to establish causal relationships in AF26. We identified the missense variant TTN-T32756I in three unrelated subjects in our multiethnic AF registry and studied this variant in an iPSC-aCM model. Remarkably, the single amino acid change in this giant protein titin results in aberrant contractility and EP remodeling without affecting the sarcomeric integrity. Although we did not observe any defects in the sarcomere, in a separate study we showed that deletion of nine amino acids including the T32756I in the same area caused significant perturbation in the sarcomere assembly37. T32756 corresponds to a conserved Threonine of the Ig139 domain (UniProt# Q8WZ42, predicted Ig155 by TITINdb2) and the replacement of a conserved hydrophilic residue (Thr) in WT with a hydrophobic residue (Ile) may cause thermal instability and destabilize the domain. It is now well recognized that point mutations may drastically disrupt the Ig domains and could possibly unfold under pathological conditions51. Such mutations have been proposed to modify titin-based passive stiffness and may lead to cardiomyopathy.

Missense mutations such as T2850I, R57C, S22P in the I-band titin have been reported to significantly destabilize Ig domains and display a higher tendency for unfolding52,53. Likewise, T32756I may unfold under the physiological settings and increase susceptibility to protein loss (haploinsufficiency) or degradation (poison peptide effect), leading to impaired sarcomere function. Furthermore, as titin acts as a crucial signaling hub that transduces mechanical forces to downstream signaling pathways within cardiomyocytes and interacts with various sarcomeric proteins and signaling molecules14,54, subtle destabilization of Ig139 domain by the T32756I could alter the biochemical signals that regulate various downstream signaling pathways involved in cardiac function and adaptation. Hence, the contractile and ion channel dysfunction caused solely by the homozygosity of T32756I in the iPSC-aCMs provide strong evidence that the AF is exclusively caused by the TTN-T32756I missense variant.

Although several studies have identified possible pathways associated with TTNtvs and AF, the pathogenic mechanisms of TTNmv-associated AF remain unclear. We show for the first time that the T32756I creates an EP substrate for AF by ion channel remodeling with an enhanced IKs, which is mediated by a titin interacting protein FHL2. FHL2 is enriched in cardiac muscle and is a multifunctional protein that regulates cardiac myocyte signaling and function55. FHL2 interacts with KCNE1 that binds to the outer face of the KCNQ1 channel pore domain and may modify the interactions between the voltage sensor, S4-S5 linker, and the pore domain to augment the channel current43. It has been suggested that FHL2 interacts with titin N2B segment, titin kinase domain and forms a complex with other proteins, such as MURF1 and MURF2, calmodulin, and Nbr1, playing a role in hypertrophy signaling and the atrophy response56,57. It also binds to the edge of titin (M-line), which is in close proximity to T32756I, and interacts with myospryn and obscurin58. It is tempting to postulate that T32756I may interfere with FHL2’s ability to interact with titin, which could then disrupt with subsequent mechano-biochemical signaling and increase FHL2’s availability for binding with the KCNQ1-KCNE1 complex. While our study provides mechanistic insights into the role of the TTN-32756I missense variant in AF, TTNmvs may be involved in the regulation of ion channel modulation and atrial rhythmicity through a multitude of interrelated signaling pathways. Further studies are needed to fully uncover the molecular basis of this regulation and its implications for AF pathophysiology and potential therapeutic interventions.

In summary, this study supports a causal relationship between a TTNmv and AF and suggests that TTNmvs may be associated with adverse outcomes in patients with AF. The study also unveils the critical role of FHL2 in modulating the atrial action potential via binding to Kv7.1-KCNE1 and enhancing the IKs in TTN-T32756I-iPSC-aCMs. Importantly, we found that the increased IKs can be rescued by FHL2 suppression. This suggests that targeting IKs or FHL2 may restore and maintain sinus rhythm and improve atrial contractility. Further study is needed to determine if TTNmvs in other regions are associated with AF by a similar mechanism. Our data indicate that AF-associated TTNmv can cause arrhythmia, provide new insight into the mechanisms underlying AF and suggest that targeting KCNQ1 or FHL2 could represent a new therapeutic strategy for improving cardiac function. Our findings may also have broader implications for the treatment of patients harboring disease-causing rare variants in sarcomeric proteins and suggest that genomic analyses that encompass TTNmvs should also consider experimental evidence in advancing our understanding of AF etiology and improving patient care in the era of precision medicine.

Methods

Study population

The UIC Multi-Ethnic AF Biorepository was established in 2015 to explore the genetic basis of AF across race and ethnicity. This study was approved by the UIC Institutional Review Board (#2015-0681). Subjects are prospectively enrolled from outpatient and inpatient sites within the University of Illinois Health (UIH) healthcare system. Subjects must have a documented history of AF by electrocardiogram, Holter/event monitor, or implantable cardiac device, with confirmation by an attending cardiologist. At time of enrollment, blood is drawn for DNA extraction and genetic analysis. Baseline demographics are obtained through provider notes in the electronic health record (EHR) and confirmed with the patient at the time of enrollment as necessary. Longitudinal clinical outcomes are followed through serial EHR review. Written informed consent was obtained from all participants. Adults 18 years of age or older at time of AF diagnosis were prospectively enrolled between August 25, 2015, and May 19, 2019. Samples from 161 subjects who identified as non-Hispanic Black (NHB) or Hispanic/Latinx (HL) and had an echocardiogram performed within 3 months of enrollment date underwent whole exome sequencing. A total of 27 subjects with congenital or rheumatic disease, severe mitral stenosis, and end stage renal disease on dialysis were excluded. Three subjects who carried a predicted loss-of-function frameshift or stop-gain TTN variant were additionally excluded. The primary exposure was the presence of one or more missense TTN variants. The primary outcome was time from initial AF diagnosis to hospitalizations with a primary diagnosis of either AF or acute decompensated HF recorded in the clinical discharge summary, captured as repeating events. Clinical data was obtained through manual review of EHR and comorbidities were defined if present in clinical notes on or prior to AF diagnosis date. Subjects were censored on date of death, last known clinical encounter (up to August 26, 2023), or at 10 years after AF diagnosis. Echocardiographic criteria for left ventricular dilatation was defined as left ventricular end diastolic diameter (LVEDD) > 2 standard deviations (SD) above sex-specific mean according to standard guidelines59. Subjects meeting echocardiographic criteria for left ventricular dilatation fulfilled clinical criteria for nonischemic dilated cardiomyopathy if left ventricular ejection fraction (LVEF) < 50% and ischemic cardiomyopathy was ruled out by coronary angiogram.

Whole exome sequencing

Samples were sequenced with support of the National Human Genome Research Institute (NHGRI) Centers for Common Disease Genetics (CCDG) program. Sequencing was performed at the Broad Institute following methods established by the National Heart, Lung, and Blood Institute (NHLBI) Trans-Omics for Precision Medicine (TOPMed) Atrial Fibrillation Study (available under TOPMed Whole Genome Sequencing Methods: Freeze 9 at https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs001062.v5.p2). Samples underwent WES using an Illumina HiSeq X system with 150bp paired-end read length, at 20x depth for at least 85% of targets Alignment to human genome reference GRCh38/hg38 was performed using BWA-MEM (Burrows-Wheeler Aligner, v0.7.15.r1140). TTN variant calls were filtered to select those with read depth of ≥20X, genotype quality ≥20, minor allele frequency (MAF) ≤1% across gnomAD subpopulations, and excluding intronic, regulatory or UTR 3’/5’ variants. Synonymous variants, homozygous reference calls, non-canonical variants, and any variants with percent spliced in (PSI) index <90 were excluded. Missense variants predicted to be deleterious were identified using REVEL score of ≥0.7, a previously proposed cutoff to predict deleterious effect in dilated cardiomyopathy60,61. Because of the uncertain role of missense TTN variants across AF and other cardiac diseases, we examined all TTN missense variants regardless of REVEL score. Variant annotation data were obtained from Ensembl Variant Effect Predictor62, downloaded January 11, 2024, and CardioDB (https://cardiodb.org/titin) for TTN region and percent spliced in (PSI)63, downloaded November 16, 2023.

Human iPSC culture and human iPSC-aCM differentiation

Human iPSC-CMs were derived from reprogrammed peripheral blood mononuclear cells (PBMCs) as previously described29. 80-90% confluent iPSC-CMs was differentiated using the Cardiomyocyte Differentiation Kit (Gibco) and guided toward the atrial subtype using all-trans RA64. The cellular population was purified through glucose starvation and lactate replacement, resulting in contracting monolayers of iPSC-aCMs. Our protocol typically yields ∼80 to 90% pure iPSC-aCMs and <6% fibroblasts based on immunostaining analysis as we have previously described29,64. Human iPSC-aCMs were then matured following dissociation and replating on fibronectin-coated plates and maintained in Cardiomyocyte Maintenance Media supplemented with T3, insulin-like growth factor-1, and dexamethasone as previously described27.

Generation of TTN-T32756I hiPSCs using CRISPR-Cas9

We edited the genome of WT iPSCs to introduce the T32756I using CRISPR-Cas9 technique29. The TTN-T32756I CRISPR-Cas9 was designed from WT allele sgRNA and two ssODN that target the TTN genomic locus: 31111-31119 and genomic DNA location: chr2:178539785-178539811 (GRCh38.p13). Amplification in the exon 352 cDNA donor RNP was constructed and electroporated into the TTN gene WT hiPSCs. Gene-editing efficiency was confirmed with next-generation sequencing (NGS). Karyotype analysis was carried out by the Cytogenetics laboratory at WiCell Research Institute Inc. Cells were collected and chromosome were evaluated using the giemsa trypsin wright (GTW) banding method. Metaphase cells were analyzed, all of which were concluded to have a normal karyotype (46, XY)

Contraction analysis and optical voltage mapping

Contractility of hiPSC-aCMs was performed with a versatile open-source software, MUSCLEMOTION65. MUSCLEMOTION is ab ImageJ plugin that utilizes a video-based system to assess contractile functions. Optical voltage mapping recordings were performed on the IonOptix system myopacer system using the fluovolt membrane potential kit (Thermo Fisher). HiPSC-aCMs were cultured in confocal dishes were incubated with Tyrode’s solution (140 mM NaCl, 4.56 mM KCl, 0.73 mM MgCl2, 10 mM HEPES, 5.0 mM dextrose, 1.25 mM CaCl2) plus 1x Fluovolt (Sigma/Aldrich) for 15-20 minutes. Cells were then rinsed with normal Tyrode’s solution and recorded APDs on the IonOptix system. These experiments were performed at the room temperature (25°C).

Electrophysiology

Whole-cell patch clamping on hiPSC-aCMs IK recordings was performed according to previously published protocols27,29. Briefly, voltage-clamps were achieved by using an Axopatch 200B amplifier controlled by pClamp10 software through an Axon Digidata 1440A. External solution for IK recordings contained: 140 mM NaCl, 4 mM KCl, 1.8 mM CaCl2, 1.2 mM MgCl2, 10 mM glucose, 10 mM HEPES, and 0.01 mM nifedipine, adjusted to pH 7.4 with NaOH. IKs recordings were isolated as 1 μM HMR-1556–sensitive current. The intracellular solution contained 100 mM potassium aspartate, 2 mM MgCl2, 20 mM KCl, 5 mM Mg-ATP, 5 mM EGTA, and 10 mM HEPES adjusted to 7.2 with KOH. IKs currents were elicited by using 3-second voltage-clamp steps to test potentials of −60 to +60 mV from holding potential of –40 mV and with 20 mV increments.

Calcium Handling

Calcium transients were measured using Fluo-4-AM (Invitrogen) dye dissolved in 2.5% Pluronic F-127 (MilliporeSigma). To get a working concentration of 5 μM, the dye solution was added to Tyrode’s solution (1 mM Ca2+). The cells were treated with Fluo-4-AM in 1 mM Ca2+ Tyrode’s solution, and then allowed to sit at room temperature for 20 minutes in the dark. Tyrode’s solution without indicators and containing 2 mM calcium was used to wash the cells. Using a 40× objective and the Zeiss LSM 710 confocal equipped with a BiG module, line scans were acquired and examined using ImageJ. The corrected minimum and maximum fluorescence values were determined by normalizing the fluorescence using a baseline background region that was unique to each cell. Fluorescence signals were normalized to basal cell fluorescence after fluo-4 loading (F0). Estimations of intracellular Ca2+ are reported as changes in ΔF/F0, where ΔF = F-F0.

Transmission electron microscopy (TEM)

iPSC-aCMs were fixed with 2.5% glutaraldehyde in 0.1M Sorenson’s Buffer for 60 minutes at room temperature. The cells are then carefully collected into a microcentrifuge tube pre-filled with same the fixation buffer. The samples were then centrifuged at 2500xg for 10 min at room temperature. The pellet was then removed and inverted with a hypodermic needle to ensure fixative solution thoroughly permeates the sample. The pellet is then allowed to incubate at room temperature to fix for an additional 60 minutes at room temperature. The fixation solution was then removed and substituted with 1% glutaraldehyde + 4% paraformaldehyde in 0.1M Sorenson’s buffer and stored in 4°C. Fixed samples were embedded in resin and microtome sections were imaged on a JEOL JEM-1400 Flash TEM.

Immunofluorescence

Cells were cultured on Matek(R) glass bottom dishes that were coated with vitronectin. The cells were then allowed to grow for a period of 2 days. The cells were then fixed with 4% paraformaldehyde (PFA) for 10 minutes and permeabilized with phosphate-buffered saline (PBS) containing 0.1% Triton X-100 for 10 minutes. After that, the cells were blocked with 3% bovine serum albumin (BSA) in PBS for 1 hour at room temperature. The staining procedure was carried out overnight at 4°C using primary antibodies that were diluted in 3% BSA in PBS. The cells were rinsed three times with PBS and then exposed to secondary antibodies (anti-mouse-FITC or anti-rabbit-PE, Santacruz, 1:1000) for one hour at room temperature. Subsequently, the cells were rinsed 3x with PBS and co-stained with 1µg/ml of DAPI. The antibodies Anti-SOX2 (Abcam, 1:200) and anti-OCT4 (Santacruz, 1:200) were used to assess the generation of iPSCs, while anti-cTnT (Invitrogen, 1:300) and anti-Kv1.5 (Alomone labs, 1:200) were utilized to confirm the differentiation of atrial cardiomyocytes. The samples were visualized using an LSM710 Meta Confocal Microscope manufactured by Zeiss.

Protein isolation, western blotting, and co-immunoprecipitation

We performed Western blots as previously described27,29. For western blots, cells on 6-well plates were washed with ice cold DPBS without Ca2+ and Mg2+, after which 250 µL of 1X RIPA with protease and phosphatase inhibitors was added per well. Lysate concentrations were measured using BCA Assay, and diluted with 4X Laemmli buffer with 10% 2-Mercaptoethanol. Per sample, 25 µg of protein was then run on an SDS-polyacrylamide gel to separate, and resolved gels were electro-transferred on 0.2 µm PVDF membranes. Membranes were blocked with 5% BSA for 1 hour, and then probed with corresponding antibodies of target proteins (anit-FHL2 antibody, Abcam#ab202584). The blots were developed using either anti-rabbit HRP or anti-mouse HRP and scanned on C280 imaging systems (Azure Biosystems). Protein signal densities were determined using ImageJ and normalized to corresponding β-actin signal densities.

For co-immunoprecipitation experiments, cells are rinsed with ice cold DPBS without Ca2+ and Mg2+, after which 500 µL of 1X RIPA with protease and phosphatase inhibitors was added each well of a 6-well plate. Lysates are sonicated on ice 3×5 seconds, then centrifuged for 10 min at 14,000xg at 4°C. 100µL of Protein A/G agarose bead slurry was added to the lysate and incubated at 4°C on a rotator for 60 minutes. The cell lysate with bead slurry was centrifuged for 10 minutes at 1000xg at 4°C and the supernatant was transferred to a fresh Eppendorf tube. 2µg of primary antibody (KCNE1 [Alomone# APC-168]: 2.5µL) was added to 500µg of precleared cell lysate, and the mixture was incubated with gentle rotation overnight at 4°C for a day. 40µL of Protein A/G bead slurry was the washed 3x with RIPA, and the precleared cell lysate with the primary antibody was added to the bead slurry, and then followed by incubation for 3 hours at 4°C. The mixture is centrifuged for 10 min at 1000xg at 4°C, and the flow through was saved to verify the immunoprecipitation. The bead pellet was rinsed with 5×500 µL RIPA and resuspended in 30µL 4X Laemmli sample buffer, vortexed, then centrifuged for 30 seconds. The sample was then boiled for 5 min, then centrifuged at 14,000xg for 5 min, representing the IP sample. 15 µL of supernatant and immunoprecipitate samples were loaded onto a 10% SDS-PAGE gel and analyzed by western blotting.

RNA-Sequencing

The RNA was isolated using the Maxwell® RSC simplyRNA Cells Kit (Promega AS1390) according to the instructions provided by the manufacturer. Library preparation was carried out using the Universal Plus mRNA-Seq kit (NuGen 0520-A01). Briefly, RNA underwent poly-A selection, enzymatic fragmentation, and generation of double-stranded cDNA using a mixture of Oligo (dT) and random priming. The cDNA was subjected to end repair, ligation of dual-index adaptors, strand selection, and 15 cycles of PCR amplification. The concentrations of the purified library were determined using the Qubit 1X dsDNA HS Assay Kit (Invitrogen Q33231). The libraries were then combined in equal amounts, considering the Qubit concentration and the average size determined by TapeStation. The pooled libraries were subsequently run on the MiniSeq instrument for index balancing. The ultimate, refined pool was measured using quantitative polymerase chain reaction (qPCR). The raw reads were aligned to the reference genome hg38 using the STAR alignment tool. The quantification of ENSEMBL gene expression was performed using FeatureCounts. The edgeR package was used to calculate normalized and differential expression statistics. To account for multiple testing, p-values were adjusted using the false discovery rate (FDR) correction method. We detected differentially expressed genes (DEGs) with a q value of less than 0.05 and a log 2 fold change of at least 2 between the Control and AF iPSC-aCMs. We conducted unsupervised hierarchical clustering of all genes that showed differential expression using the Euclidean distance and complete linkage method. Additionally, we created volcano plots using the R programming language. The up- and down-regulated genes were analyzed individually using the DAVID functional annotation tool against the Gene Ontology Biological Process (GO BP) database.

qPCR

Total RNA was isolated from hiPSC-aCMs using TRIzol reagent (Invitrogen), following the manufacturer’s instructions to ensure the extraction of high-quality RNA. The concentration and purity of the isolated RNA were carefully assessed using a NanoDrop 2000 spectrophotometer (Thermo Fisher Scientific), with 1 µg of total RNA utilized for each reverse transcription reaction. Reverse transcription to synthesize cDNA was conducted using SuperScript III Reverse Transcriptase (Thermo Fisher Scientific). For the analysis, specific assays and primers were selected for target gene FHL2 (Forward sequence: GTGGTGTGCTTTGAGACCCTGT, Reverse sequence: GAGCAGTGGAAACAGGCTTCATG) with glyceraldehyde 3-phosphate dehydrogenase (GAPDH) serving as the normalization reference gene. qPCR reactions were performed on an ABI QuantStudio 5 system (Applied Biosystems), using SYBR Green PCR Master Mix to accurately detect and quantify PCR amplification products. Relative expression levels of the target genes were calculated employing the ΔΔCt method, by the quantification of gene expression changes in the experimental samples relative to control. For the ΔCT Calculation, the cycling time (CT) value of the target gene was subtracted from the CT value of GAPDH in the same sample using ΔCT=CTtarget gene − CTreference gene. The ΔΔCT value was then calculated using ΔΔCT= ΔCTExperimental – ΔCTControl. The relative expression for the gene was in turn calculated using Relative gene expression= 2−ΔΔCT.

siRNA experiments

FHL2-specific siRNAs (#SR301594) or scrambled siRNAs were applied to mature hiPSC-aCMs using Lipofectamine RNAiMAX (Invitrogen, Carlsbad, CA). Opti-MEM medium was used to dilute stock solutions of lipofectamine and siRNA, each prepared at a concentration of 10 µM. The siRNA-lipid complex was then made by mixing these solutions in a 1:1 ratio and incubated for five minutes. After adding this complex to the cells dropwise, the media was changed, and the cells were allowed to incubate for a further two days.

Data Analysis and Statistics

For clinical data, categorical variables are represented as count and percentage (%) and tested by Fisher’s exact test. Continuous variables are reported as mean (standard deviation [SD]) or median and interquartile range (IQR) where specified and tested with Mann-Whitney U and Kruskal-Wallis tests. Ordinal variables were tested with Kruskal-Wallis test. Univariable and multivariable Cox proportional hazards models were used to evaluate risk of the primary outcome as recurring events, related to presence of TTN missense variants. A partially adjusted model was first tested using covariates of age and sex, followed by a fully adjusted model which additionally incorporated covariates of ethnicity and baseline LVEF <50%. Interaction between TTN missense variants and LVEF was additionally evaluated. To identify whether associations were limited to variants with predicted deleterious effect, analysis was repeated stratifying subjects according to presence and predicted impact of TTNmv: presence of one or more predicted deleterious (REVEL≥0.7) variants, presence of predicted benign (REVEL <0.7) variants only, or no TTNmv present. Additional sensitivity analyses were performed excluding subjects with multiple TTNmvs and with nonischemic dilated cardiomyopathy. Analysis was performed in R (version 4.2.1, packages: arsenal, ggplot2, ggsurvfit, gtsummary, reda, survival).

Experiments were performed at least two times (biological replicates) to ensure reproducibility. Unless otherwise noted, experimental data on hiPSCs are shown as mean ± SD. *P < 0.05, **P < 0.01, ***P < 0.001, ****P < 0.0001 indicate significance, while P > 0.05 is regarded as non-significant. Statistical analyses include unpaired nonparametric and 2-tailed Mann-Whitney U test for data with normal distribution, and either 1-way or 2-way ANOVA with post hoc Bonferroni’s corrections for several groups. The first and third quartiles of a median are used to express skewed data. Fisher’s Exact test is used to compare categorical data, and unpaired Student’s t-test or ANOVA are applied to assess continuous variables.

eLife Assessment

The study presents valuable findings regarding the incidence and clinical impact of a mutation in a cardiac muscle protein and its association with the development of atrial fibrillation. The authors provide some convincing evidence of electrophysiological disturbances in cells with this mutation which would be of interest to cellular electrophysiologists. However, evidence supporting the conclusion that this mutation causes atrial fibrillation would benefit from more rigorous electrophysiologic approaches.

Reviewer #1 (Public review):

Summary:

Pavel et al. analyzed a cohort of atrial fibrillation (AF) patients from the University of Illinois at Chicago, identifying TTN truncating variants (TTNtvs) and TTN missense variants (TTNmvs). They reported a rare TTN missense variant (T32756I) associated with adverse clinical outcomes in AF patients. To investigate its functional significance, the authors modeled the TTN-T32756I variant using human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs). They demonstrated that mutant cells exhibit aberrant contractility, increased activity of the cardiac potassium channel KCNQ1 (Kv7.1), and dysregulated calcium homeostasis. Interestingly, these effects occurred without compromising sarcomeric integrity. The study further identified increased binding of the titin-binding protein Four-and-a-Half Lim domains 2 (FHL2) with KCNQ1 and its modulatory subunit KCNE1 in the TTN-T32756I iPSC-aCMs.

Strengths:

This work has translational potential, suggesting that targeting KCNQ1 or FHL2 could represent a novel therapeutic strategy for improving cardiac function. The findings may also have broader implications for treating patients with rare, disease-causing variants in sarcomeric proteins and underscore the importance of integrating genomic analysis with experimental evidence to advance AF research and precision medicine.

Weaknesses:

(1) Variant Identification: It is unclear how the TTN missense variant (T32756I) was identified using REVEL, as none of the patients' parents reportedly carried the mutation or exhibited AF symptoms. Are there other TTN variants identified in the three patients carrying TTN-T32756I? Clarification on this point is necessary.

(2) Patient-Specific iPSC Lines: Since the TTN-T32756I variant was modeled using only one healthy iPSC line, it is unclear whether patient-specific iPSC-derived atrial cardiomyocytes would exhibit similar AF-related phenotypes. This limitation should be addressed.

(3) Hypertension as a Confounding Factor: The three patients carrying TTN-T32756I also have hypertension. Could the hypertension associated with this variant contribute secondarily to AF? The authors should discuss or rule out this possibility.

(4) FHL2 and KCNQ1-KCNE1 Interaction: Immunostaining data demonstrating the colocalization of FHL2 with the KCNQ1-KCNE1 (MinK) complex in TTN-T32756I iPSC-aCMs are needed to strengthen the mechanistic findings.

(5) Functional Characterization of FHL2-KCNQ1-KCNE1 Interaction: Additional functional assays are necessary to characterize the interaction between FHL2 and the KCNQ1-KCNE1 complex in TTN-T32756I iPSC-aCMs to further validate the proposed mechanism.

Reviewer #2 (Public review):

Summary:

The authors present data from a single-center cohort of African-American and Hispanic/Latinx individuals with atrial fibrillation (AF). This study provides insight into the incidences and clinical impact of missense variants in the Titin (TTN) gene in this population. In addition, the authors identified a single amino acid TTN missense variant (TTN-T32756I) that was further studied using human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs). These studies demonstrated that the Four-and-a-Half Lim domains 2 (FHL2), has increased binding with KCNQ1 and its modulatory subunit KCNE1 in the TTN-T32756I-iPSC-aCMs, enhancing the slow delayed rectifier potassium current (Iks) and is a potential mechanism for atrial fibrillation. Finally, the authors demonstrate that suppression of FHL2 could normalize the Iks current.

Strengths:

The strengths of this manuscript/study are listed below:

(1) This study includes a previously underrepresented population in the study of the genetic and mechanistic basis of AF.

(2) The authors utilize current state-of-the-art methods to investigate the pathogenicity of a specific TTN missense variant identified in this underrepresented patient population.

(3) The findings of this study identify a potential therapeutic for treating atrial fibrillation.

Weaknesses:

(1) The authors do not include a non-AF group when evaluating the incidence and clinical significance of TTN missense variants in AF patients.

(2) The authors do not provide evidence that TTN-T32756I-iPSC-aCMs are arrhythmogenic only that there is an increase in the Iks current and associated action potential changes. More specifically, the authors report "compared to the WT, TTN-T32756I-iPSC-aCMs exhibited increased arrhythmic frequency" yet is it is unclear what they are referring to by "arrhythmic frequency".

(3) There seem to be discrepancies regarding the impact of the TTN-T32756I variant on mechanical function. Specifically, the authors report "both reduced contraction and abnormal relaxation in TTN-T32756I-iPSC-aCMs" yet, separately report "the contraction amplitude of the mutant was also increased . . . suggesting an increased contractile force by the TTN-T32756I-iPSC-aCMs and TTN-T32756I-iPSC-CMs exhibited similar calcium transient amplitudes as the WT."

Reviewer #3 (Public review):

Summary:

The authors describe the abnormal contractile function and cellular electrophysiology in an iPSC model of atrial myocytes with a titin missense variant. They provide contractility data by sarcomere length imaging, calcium imaging, and voltage clamp of the repolarizing current iKs. While each of the findings is separately interesting, the paper comes across as too descriptive because there is no merging of the data to support a cohesive mechanistic story/statement, especially from the electrophysiological standpoint. There is definitely not enough support for the title "A Titin Missense Variant Causes Atrial Fibrillation", since there is no strong causative evidence at all. There is some interesting clinical data regarding the variant of interest and its association with HF hospitalization, which may lead to future important discoveries regarding atrial fibrillation.

Strengths:

The manuscript is well written and there is a wide range of experimental techniques to probe this atrial fibrillation model.

Weaknesses:

(1) While the clinical data is interesting, it is extremely important to rule out heart failure with preserved EF as a confounder. HFpEF leads to AF due to increased atrial remodeling, so the fact that patients with this missense variant have increased HF hospitalizations does not necessarily directly support the variant as causative of AF. It could be that the variant is actually associated directly with HFpEF instead, and this needs to be addressed and corrected in the analyses.

(2) All of the contractility and electrophysiologic data should be done with pacing at the same rate in both control and missense variant groups, to control for the effect of cycle length on APD and calcium loading. A claim of shorter APD cannot be claimed when the firing rate of one set of cells is much faster than the other, since shorter APD is to be expected with a faster rate. Similarly, contractility is affected by diastolic interval because of the influence of SR calcium content on the myocyte power stroke. So the cells need to be paced at the same rate in the IonOptix for any direct comparison of contractility. The authors should familiarize themselves with the concept of electrical restitution.

(3) It is interesting that the firing rate of the myocytes is faster with the missense variant. This should lead to a hypothesis and investigation of abnormal automaticity or triggered activity, which may also explain the increased contractility since all these mechanisms are related to the calcium clock and calcium loading of the SR. See #2 above for suggestions on how to adequately probe calcium handling. Such an investigation into impulse initiation mechanisms would be very powerful in supporting the primary statement of the paper since these are actual mechanisms thought to cause AF.

(4) The claim of shortened APD without correcting for cycle length is problematic. However, the general concept of linking shortened APD in isolated cells alone to AF causation is more problematic. To have a setup for reentry, there must be a gradient of APD from short to long, and this can only be demonstrated at the tissue level, not really at the cellular level, so reentry should not be invoked here. If shortened APD is demonstrated with correction of the cycle length problem, restitution curves can be made showing APD shortening at different cycle lengths. If restitution is abnormal (i.e. the APD does not shorten normally in relation to the diastolic interval), this may lead to triggered activity which is an arrhythmogenic mechanism. This would also tie in well with the finding of abnormally elevated iKs current since iKs is a repolarizing current directly responsible for restitution.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

Pavel et al. analyzed a cohort of atrial fibrillation (AF) patients from the University of Illinois at Chicago, identifying TTN truncating variants (TTNtvs) and TTN missense variants (TTNmvs). They reported a rare TTN missense variant (T32756I) associated with adverse clinical outcomes in AF patients. To investigate its functional significance, the authors modeled the TTN-T32756I variant using human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs). They demonstrated that mutant cells exhibit aberrant contractility, increased activity of the cardiac potassium channel KCNQ1 (Kv7.1), and dysregulated calcium homeostasis. Interestingly, these effects occurred without compromising sarcomeric integrity. The study further identified increased binding of the titin-binding protein Four-and-a-Half Lim domains 2 (FHL2) with KCNQ1 and its modulatory subunit KCNE1 in the TTN-T32756I iPSCaCMs.

Strengths:

This work has translational potential, suggesting that targeting KCNQ1 or FHL2 could represent a novel therapeutic strategy for improving cardiac function. The findings may also have broader implications for treating patients with rare, disease-causing variants in sarcomeric proteins and underscore the importance of integrating genomic analysis with experimental evidence to advance AF research and precision medicine.

Weaknesses

(1) Variant Identification: It is unclear how the TTN missense variant (T32756I) was identified using REVEL, as none of the patients' parents reportedly carried the mutation or exhibited AF symptoms. Are there other TTN variants identified in the three patients carrying TTN-T32756I? Clarification on this point is necessary.

We thank the reviewer for their insightful comment. We have now clarified these in the method section.

Line 484-491: “The TTN-T32756I variant (REVEL Score: 0.58758, Supplementary Table 1) was prioritized due to its occurrence in multiple unrelated individuals within our clinical AF cohort, despite no reported family history of AF in affected individuals. While no parental inheritance was observed, the possibility of de novo origin cannot be excluded. Furthermore, this variant is located within a region overlapping a deletion mutation recently shown to cause AF in a zebrafish model, supporting its potential pathogenicity [37]. Notably, the affected individuals did not carry additional loss-of-function TTN variants.”

(2) Patient-Specific iPSC Lines: Since the TTN-T32756I variant was modeled using only one healthy iPSC line, it is unclear whether patient-specific iPSC-derived atrial cardiomyocytes would exhibit similar AF-related phenotypes. This limitation should be addressed.

We have now acknowledged this limitation in the revised manuscript.

Line 505-509: “Due to the patients' unavailability of peripheral blood mononuclear cells (PBMCs), we utilized a healthy iPSC line and introduced the TTN-T32756I variant using CRISPR/Cas9 genome editing. This approach ensures an isogenic background, thereby minimizing genetic variability and providing a controlled system to study the direct effects of the mutation.”

(3) Hypertension as a Confounding Factor: The three patients carrying TTN-T32756I also have hypertension. Could the hypertension associated with this variant contribute secondarily to AF? The authors should discuss or rule out this possibility.

We have now explicitly discussed this in the revised manuscript.

Line 362-367: “Hypertension is a common comorbidity in patients with AF and could contribute to disease progression. However, all three individuals carrying TTN-T32756I exhibited earlyonset AF (onset before 66 years), with one case occurring as early as 36 years. This suggests a potential two-hit mechanism, where genetic predisposition and comorbidities influence disease risk. Importantly, our iPSC model isolates the genetic effects of TTN-T32756I from other factors, supporting a direct pathogenic role.”

(4) FHL2 and KCNQ1-KCNE1 Interaction: Immunostaining data demonstrating the colocalization of FHL2 with the KCNQ1-KCNE1 (MinK) complex in TTN-T32756I iPSC-aCMs are needed to strengthen the mechanistic findings.

We thank the reviewer for this insightful suggestion. We agree that additional immunostaining data would further strengthen the evidence for FHL2 colocalization with the KCNQ1-KCNE1 complex in TTN-T32756I iPSC-aCMs. In line with this, we have expanded our analysis to include both co-immunoprecipitation and confocal microscopy. As described in the revised manuscript (Lines 282–287), the colocalization between KCNE1 and FHL2 was increased by approximately threefold in TTN-T32756I iPSC-aCMs compared with WT, supporting an enhanced interaction between these proteins (Figure 5A, Supplementary Figure 6). We are generating additional immunostaining data to validate and extend these findings, and we will incorporate them into the revised submission to further substantiate the mechanistic link proposed.

Line 282-287: “…..if TTN-T32756I increases Iks by modulating the interaction between KCNQ1KCNE1 and FHL2, we performed co-immunoprecipitation studies and confocal microscopy in both WT and TTN-T32756I-iPSC-aCMs. The co-localization between KCNE1 and FHL2 increased ~3 fold in TTN-T32756I-iPSC-aCMs, suggesting an increased interaction between them (Figure 5A, Supplementary Figure 7).”

(5) Functional Characterization of FHL2-KCNQ1-KCNE1 Interaction: To further validate the proposed mechanism, additional functional assays are necessary to characterize the interaction between FHL2 and the KCNQ1-KCNE1 complex in TTN-T32756I iPSC-aCMs.

We thank the reviewer for this valuable suggestion. We agree that additional functional assays would provide further validation of the proposed mechanism. However, we believe such in-depth characterization warrants a dedicated follow-up study and is beyond the scope of the current revision. In this work, our primary objective is to establish that the TTN missense variant can exert a detrimental effect and serve as a substrate for AF.

Line 418-419: “Further study is needed to validate the proposed mechanism and determine if TTNmvs in other regions are associated with AF by a similar process.”

Reviewer #2 (Public review):

Summary:

The authors present data from a single-center cohort of African-American and Hispanic/Latinx individuals with atrial fibrillation (AF). This study provides insight into the incidences and clinical impact of missense variants in this population in the Titin (TTN) gene. In addition, the authors identified a single amino acid TTN missense variant (TTN-T32756I) that was further studied using human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs). These studies demonstrated that the Four-and-a-Half Lim domains 2 (FHL2) has increased binding with KCNQ1 and its modulatory subunit KCNE1 in the TTN-T32756I-iPSCaCMs, enhancing the slow delayed rectifier potassium current (Iks) and is a potential mechanism for atrial fibrillation. Finally, the authors demonstrate that suppression of FHL2 could normalize the Iks current.

Strengths:

The strengths of this manuscript/study are listed below:

(1) This study includes a previously underrepresented population in the study of the genetic and mechanistic basis of AF.

(2) The authors utilize current state-of-the-art methods to investigate the pathogenicity of a specific TTN missense variant identified in this underrepresented patient population.

(3) The findings of this study identify a potential therapeutic for treating atrial fibrillation.

Weaknesses:

(1) The authors do not include a non-AF group when evaluating the incidence and clinical significance of TTN missense variants in AF patients.

We appreciate the reviewer’s comment and acknowledge the limitation of not including a non-AF control group in our clinical analysis. As noted in the revised manuscript (Lines 347–353), our cohort was derived from a single-center registry of individuals with AF and therefore lacks a matched non-AF control population for direct comparison of TTN missense variant incidence. We agree that future studies incorporating larger, multiethnic validation cohorts with both AF and non-AF individuals, as well as evaluating AF-specific measures such as arrhythmia burden and treatment response, will be essential to fully elucidate the clinical significance of TTN missense variants in AF.

Line 347-353: “Our cohort is derived from a single-center multi-ethnic registry of individuals with AF and lacks a matched cohort of non-AF controls to compare the incidence of TTN missense variants. Further study exploring these associations in mult-ethnic, larger validation cohorts that include both AF and non-AF individuals and examining AF-specific measures such as arrhythmia burden or treatment response will be necessary to fully understand the clinical importance of TTNmvs in AF.”

(2) The authors do not provide evidence that TTN-T32756I-iPSC-aCMs are arrhythmogenic, only that there is an increase in the Iks current and associated action potential changes. More specifically, the authors report that "compared to the WT, TTN-T32756I-iPSC-aCMs exhibited increased arrhythmic frequency," yet it is unclear what they are referring to by "arrhythmic frequency."

We thank the reviewer for this important point and for highlighting the need for clarification. In our study, the term “arrhythmic frequency” was intended to describe the increased spontaneous beating rate, irregular action potential patterns, and abnormal calcium handling observed in TTN-T32756I iPSC-aCMs compared with WT. These findings support the concept that the AF-associated TTN-T32756I variant promotes ion channel remodeling and perturbs excitation–contraction coupling, thereby creating a potential arrhythmogenic substrate for AF. To avoid ambiguity, we have removed the term “arrhythmic frequency” and revised the text for clarity and precision (Lines 222–223).

Lines 222-223: “Compared to the WT, TTN-T32756I-iPSC-aCMs exhibited increased frequency along with a significant reduction of the time to 50% and 90% decline of calcium transients (Figure 3G-I, Supplementary Figure 4F).”

(3) There seem to be discrepancies regarding the impact of the TTN-T32756I variant on mechanical function. Specifically, the authors report "both reduced contraction and abnormal relaxation in TTN-T32756I-iPSC-aCMs" yet, separately report "the contraction amplitude of the mutant was also increased . . . suggesting an increased contractile force by the TTN-T32756IiPSC-aCMs and TTN-T32756I-iPSC-CMs exhibited similar calcium transient amplitudes as the WT."

We thank the reviewer for highlighting this critical point and apologize for the lack of clarity. We intended to distinguish between changes in contractile force and contractile dynamics. Specifically, the increased contraction amplitude observed in TTN-T32756I iPSCaCMs reflects enhanced contractile force, whereas the reduced contraction duration and impaired relaxation reflect abnormalities in contractile kinetics. Together, these findings indicate that the TTN-T32756I variant alters both the strength and the temporal dynamics of contraction, consistent with dysfunctional mechanical performance. We have revised the text accordingly to more accurately convey these results (Lines 187–192).

Lines 187-192: “Compared to WT, the beating frequency of the TTN-T32756I-iPSC-aCMs was significantly increased (52 ± 7.8 vs. 98 ± 7.5 beats per min, P=0.001; Figure 2C) coupled with the reduction of the contraction duration (456.5 ± 61.45 vs 262.9 ± 48.16 msec, P=0.032; Figure 2D), the peak-to-peak time (1529 ± 195.5 vs 636.6 ± 135.8 msec, P=0.004; Supplementary Figure 3B), and the relaxation (281.5 ± 42.95 vs 79.40 ± 21.14 msec, P=0.003; Supplementary Figure 3A).”

Reviewer #3 (Public review):

Summary:

The authors describe the abnormal contractile function and cellular electrophysiology in an iPSC model of atrial myocytes with a titin missense variant. They provide contractility data by sarcomere length imaging, calcium imaging, and voltage clamp of the repolarizing current iKs. While each of the findings is interesting, the paper comes across as too descriptive because there is no data merging to support a cohesive mechanistic story/statement, especially from the electrophysiological standpoint. There is not enough support for the title "A Titin Missense Variant Causes Atrial Fibrillation", since there is no strong causative evidence. There is some interesting clinical data regarding the variant of interest and its association with HF hospitalization, which may lead to future important discoveries regarding atrial fibrillation.

Strengths:

The manuscript is well written, and a wide range of experimental techniques are used to probe this atrial fibrillation model.

Weaknesses

(1) While the clinical data is interesting, it is essential to rule out heart failure with preserved EF as a confounder. HFpEF leads to AF due to increased atrial remodeling, so the fact that patients with this missense variant have increased HF hospitalizations does not necessarily directly support the variant as causative of AF. It could be that the variant is associated directly with HFpEF instead, and this needs to be addressed and corrected in the analyses.

We appreciate the reviewer’s insightful comment and agree that HFpEF-related atrial remodeling could represent a potential confounder in the association between TTN missense variants and AF. The primary aim of our clinical analysis was to assess the potential significance of TTNmv in AF, recognizing the inherent limitations of retrospective observational data in establishing causality. To complement this, our in vitro studies were specifically designed to demonstrate that TTNmv can alter the electrophysiological substrate, thereby predisposing to AF independent of clinical comorbidities.

While HFpEF is an important consideration, to our knowledge, no existing literature directly implicates TTNmv in HFpEF pathogenesis. In contrast, loss-of-function TTN variants are more commonly associated with HFrEF and dilated cardiomyopathy, and even these associations remain an area of active debate. To address potential confounding in our cohort, we adjusted for reduced ejection fraction in multivariable analyses of clinical outcomes. Additionally, we performed a sensitivity analysis excluding patients with nonischemic dilated cardiomyopathy (Supplementary Table 6). Together, these approaches mitigate the potential impact of heart failure subtypes on our findings, while our mechanistic studies strengthen the argument that TTNmv may contribute directly to AF susceptibility.

(2) All contractility and electrophysiologic data should be done with pacing at the same rate in both control and missense variant groups, to control for the effect of cycle length on APD and calcium loading. A shorter APD cannot be claimed when the firing rate of one set of cells is much faster than the other, since shorter APD is to be expected with a quicker rate. Similarly, contractility is affected by diastolic interval because of the influence of SR calcium content on the myocyte power stroke. So the cells need to be paced at the same rate in the IonOptix for any direct comparison of contractility. The authors should familiarize themselves with the concept of electrical restitution.

We thank the reviewer for this crucial technical comment. iPSC-derived cardiomyocytes (iPSC-CMs) are known to exhibit spontaneous automaticity due to the presence of pacemaker-like currents and reduced IK1, which enables interrogation of their intrinsic electrophysiological properties and disease-relevant remodeling. In our study, we leveraged this feature to test the hypothesis that TTN missense variants alter electrophysiological properties through ion channel remodeling. That said, we fully agree with the reviewer that pacing iPSCCMs at a controlled cycle length is essential for minimizing rate-dependent effects on APD, calcium handling, and contractility, and would improve the interpretability of group comparisons. While iPSC-CMs with matched genetic backgrounds are expected to display broadly comparable electrophysiological profiles, biological and technical variability can influence spontaneous beating rates, thereby confounding direct comparisons. To address this, we have incorporated pacing protocols into our revised experimental design to ensure that APD and contractility measurements are obtained under identical cycle lengths, consistent with the concept of electrical restitution.

(3) It is interesting that the firing rate of the myocytes is faster with the missense variant. This should lead to a hypothesis and investigation of abnormal automaticity or triggered activity, which may also explain the increased contractility since all these mechanisms are related to the SR's calcium clock and calcium loading. See #2 above for suggestions on how to probe calcium handling adequately. Such an investigation into impulse initiation mechanisms would be compelling in supporting the primary statement of the paper since these are actual mechanisms thought to cause AF.

We thank the reviewer for this insightful suggestion. We agree that the faster firing rate observed in TTN-T32756I iPSC-aCMs raises the possibility of abnormal automaticity or triggered activity, both of which are highly relevant to AF pathophysiology. As these mechanisms are tightly coupled to calcium handling and the SR calcium clock, further probing of calcium cycling abnormalities would provide valuable mechanistic insights. While this level of investigation is beyond the scope of the current study, we view it as a compelling future direction that could directly link TTN missense variants to impulse initiation abnormalities contributing to AF.

(4) The claim of shortened APD without correcting for cycle length is problematic. However, linking shortened APD in isolated cells alone to AF causation is more complicated. To have a setup for reentry, there must be a gradient of APD from short to long, and this can only be demonstrated at the tissue level, not at the cellular level, so reentry should not be invoked here. If shortened APD is demonstrated with correction of the cycle length problem, restitution curves can be made showing APD shortening at different cycle lengths. If restitution is abnormal (i.e. the APD does not shorten normally in relation to the diastolic interval), this may lead to triggered activity which is an arrhythmogenic mechanism. This would also tie in well with the finding of abnormally elevated iKs current since iKs is a repolarizing current directly responsible for restitution.

We thank the reviewer for this necessary clarification. We agree that isolated cell studies cannot directly demonstrate reentrant circuits and that reentry should not be inferred solely from cellular APD data. Our observation of shortened APD and abnormal beating patterns in TTN-T32756I iPSC-aCMs suggests ion channel remodeling that may predispose to arrhythmogenic conditions. Still, we recognize that tissue-level gradients of APD are required to establish reentry as a mechanism. Accordingly, we have removed mention of “the reentrant mechanism” from the revised manuscript and limited our interpretation to the cellular findings. Future studies incorporating pacing protocols and restitution curve analyses will be valuable in determining whether abnormal APD restitution and elevated IKs contribute to triggered activity, thereby providing a more direct mechanistic link to AF (Lines 101–105).

Lines 101-105: “Our study showed that the TTN-T32756I iPSC-aCMs exhibited a striking AF-like EP phenotype in vitro, and transcriptomic analyses revealed that the TTNmv increases the activity of the FHL2, which then modulates the slow delayed rectifier potassium current (IKs) to cause AF.”

Reviewer #1 (Recommendations for the authors):

Electrophysiological Phenotype in Ventricular CMs: Has the iPSC line carrying TTN-T32756I been differentiated into ventricular cardiomyocytes (iPSC-vCMs)? The reported cellular phenotype in iPSC-aCMs does not seem to specifically reflect an AF phenotype. Does the variant produce similar electrophysiological alterations in iPSC-vCMs?

We thank the reviewer for this thoughtful comment. To date, we have not differentiated the TTN-T32756I iPSC line into ventricular cardiomyocytes (iPSC-vCMs). Our current work focuses on iPSC-aCMs, where we demonstrate that the AF-associated TTNT32756I variant induces ion channel remodeling and abnormal beating patterns, thereby creating a potential arrhythmogenic substrate relevant to AF. We agree that investigating whether this variant produces similar or distinct electrophysiological alterations in iPSC-vCMs would provide essential insights into chamber-specific effects and broaden our mechanistic understanding. We have acknowledged this as a future direction in the revised manuscript (Lines 422–425).

Lines 422-425: “While we have not yet explored the effect of TTN-T32756I in iPSC-derived ventricular cardiomyocytes, it would be interesting to investigate whether this variant produces similar or distinct electrophysiological alterations in the ventricular cardiomyocytes.”

Version 2

Abstract

Rare and common genetic variants contribute to the risk of atrial fibrillation (AF). Although ion channels were among the first AF candidate genes identified, rare loss-of-function variants in structural genes, such as TTN, have also been implicated in AF pathogenesis, partly through the development of atrial myopathy; however, the underlying mechanisms are poorly understood. While TTN truncating variants (TTNtvs) have been causally linked to arrhythmia and cardiomyopathy syndromes, the role of missense variants (mvs) remains unclear. We report that rare TTNmvs are associated with adverse clinical outcomes in AF patients, and we have identified a mechanism by which a TTNmv (T32756I) causes AF. Modeling the TTN-T32756I variant using human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs) revealed that the mutant cells display aberrant contractility, increased activity of a cardiac potassium channel (KCNQ1, Kv7.1), and dysregulated calcium homeostasis without compromising the sarcomeric integrity of the atrial cardiomyocytes. We also show that a titin-binding protein, the Four-and-a-Half Lim domains 2 (FHL2), has increased binding with KCNQ1 and its modulatory subunit KCNE1 in the TTN-T32756I-iPSC-aCMs, enhancing the slow delayed rectifier potassium current (IKs). Suppression of FHL2 in mutant iPSC-aCMs normalized the IKs, supporting FHL2 as an IKs modulator. Our findings demonstrate that a single amino acid change in titin not only affects function but also causes ion channel remodeling and AF. These findings underscore the importance of high-throughput screening to assess the pathogenicity of TTNmvs and establish a mechanistic connection between titin, potassium ion channels, and sarcomeric proteins, which may represent a novel therapeutic target.

Introduction

Atrial fibrillation (AF), the most prevalent cardiac arrhythmia, affects more than 60 million people worldwide and is associated with increased risk for stroke, heart failure, and dementia, justifying it as a major public healthcare burden1,2. AF is characterized by irregular and often abnormally fast heart rates3,4. Over the last decade, tremendous progress has been made in understanding the genetic architecture of AF5–7. Genome-wide association studies have identified over 140 common loci associated with AF, while family-based studies have implicated rare variants primarily encoding ion channels8,9. Although AF has been traditionally classified as a ‘channelopathy’, variants in myocardial sarcomeric proteins such as titin have been increasingly associated with familial or early-onset AF10,11.

The TTN gene encodes for a massive myofilament (∼4200 kDa), titin, which stretches along the Z-disk (N-terminus) to the M-band (C-terminus) region of the sarcomere12,13. Titin serves as a molecular scaffold for other muscle proteins, participates in downstream signaling, and provides passive tension to cardiac muscle14,15. Due to its large size, the rate of genetic variation in TTN is high, including both truncating (TTNtvs) and missense variants (TTNmvs)16,17. The prevalence of rare TTNtvs and TTNmvs in the general population is 2% and 5.7%, respectively18–20. While TTNtvs are the most common genetic cause (10-20%) of dilated cardiomyopathy (DCM), TTNmvs are often disregarded in clinical practice, fulfilling a stand-alone benign criterion in DCM-specific variant interpretation frameworks based on the American College of Medical Genetics (ACMG) recommendations21,22. Yet, TTNmvs may have the potential to confer disease, as a recent study in two families revealed that TTNmvs in a conserved cysteine of TTN can cause DCM23. Another study demonstrated segregation of a novel TTNmv among five individuals with atrioventricular block in a Chinese family, suggesting that TTNmvs may also be implicated in arrhythmia syndromes24. To date, a relationship between rare TTNmvs and AF has not been explored at either clinical or mechanistic levels. This has potentially heightened importance across racial and ethnic groups, in whom the likelihood of detecting variants of uncertain significance (which are predominantly TTNmvs), is higher than in individuals of European descent25.

Despite the clinical importance, the pharmacological therapy of AF is limited in part because of the incomplete understanding of the myocardial substrate for AF as human atrial tissue is rarely available and the limitations of existing in vitro and in vivo models26. However, human induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-CMs) not only possess the complex array of cardiac ion channels that make up the atrial action potential (AP) but also hold great promise for modeling AF27,28. Modeling patient-specific mutations associated with familial AF using mature human iPSC-CMs offers a powerful, naturally integrated system with distinct advantages over animal models and heterologous expression systems29. Recent reports using human iPSC-CM models have elucidated the pathophysiological mechanisms of DCM30, hypertrophic cardiomyopathy31, Brugada syndrome32, long QT syndrome33, and AF27,29,34. Human iPSC-CMs also have the potential to uncover the molecular mechanisms of AF as they can be easily modified to contain the precise genetic background of the individual patient35–37.

To examine the potential role of TTNmvs in AF, we examined the prevalence of rare TTNmvs in a single-center cohort of African-American and Hispanic/Latinx individuals with AF who underwent whole exome sequencing, compared clinical characteristics between TTNmv carriers and non-carriers, and evaluated whether TTNmvs were associated with increased risk of AF or heart failure (HF)-related hospitalizations. To explore whether the TTNmvs may be mechanistically linked to the development of AF, we introduced a rare TTNmv (TTN-T32756I) into human iPSC-aCMs using clustered regularly interspaced short palindromic repeats-associated 9 (CRISPR-Cas9). We performed electrophysiological (EP), pharmacological, and mechanistic analyses to elucidate the mechanisms by which the TTNmv causes AF. Our study showed that the TTN-T32756I iPSC-aCMs exhibited a striking AF-like EP phenotype in vitro, and transcriptomic analyses revealed that the TTNmv increases the activity of the Four-and-a-half LIM domain protein 2 (FHL2), which then modulates the slow delayed rectifier potassium current (IKs) to cause AF.

Results

TTNmvs are associated with increased hospitalization risk in a multiethnic cohort with AF

Clinical characteristics

A total of 131 subjects were included (mean [SD] age at AF diagnosis 63.5 [13.8] years, 70 [53.4%] male, 38 [29%] Hispanic/Latinx, 93 [71%] non-Hispanic Black; Table 1). We identified 138 TTNmvs, most commonly in the TTN A-band (108 [78.8]%), followed by the I-band (20 [14.6%]), M-band (2 [1.5%]), near the Z-disk (6 [4.4%], and in the Z-disk (2 [1.5%]) (Figure 1A, Supplementary Table 1). Based on the REVEL in silico score, 52 (37.7%) TTNmvs variants were predicted to be deleterious. A total of 77 (58.8%) subjects carried a TTNmv, with 43 (32.8%) carrying a predicted deleterious variant. Carriers of a TTNmv had a higher QTc interval on electrocardiogram (ECG) closest to AF diagnosis (mean [SD] 466.5 [42.3] vs. 449.6 [37.7] ms, P=0.027); there were otherwise no other significant clinical or demographic differences between variant carriers and non-carriers. When stratifying by predicted deleterious TTNmv only (Supplementary Table 2), a higher proportion of variant carriers were non-Hispanic Black (36 [83.7%] vs. 57 [64.8%], P=0.026). On index ECG, carriers additionally had higher ventricular rate (mean [SD] 103.7 [31.4] vs. 90.9 [27.4] beats/min, P=0.022), QTc interval (mean [SD] 470.6 [44.0] vs. 453.9 [38.7] ms, P=0.035), and LVEDD on index echocardiogram (mean [SD] 49.8 [8.0] vs. 45.6 [9.2] mm, P=0.021). Seventeen (15.2%) subjects met criteria for left ventricular (LV) dilatation, with a higher proportion of LV dilatation in TTNmv carriers but without a significant difference compared to non-carriers (13 [20.3%] vs. 4 [8.3%], P=0.111). Twelve subjects (9.2%) met criteria for a clinical diagnosis of nonischemic DCM based on a left ventricular ejection fraction (LVEF) <50%, presence of LV dilatation, and confirmation of non-severe coronary artery disease by coronary angiography. Of those, 8 subjects carried TTNmvs of which 6 were predicted to be deleterious (Supplementary Table 3).

TTNmv prevalence and association with hospitalization in a multiethnic atrial fibrillation (AF) cohort.

(A) Distribution of TTNmvs in a multiethnic AF cohort based on amino acid position in the TTN gene, stratified by REVEL in silico score for prediction of deleterious effect, defined by REVEL ≥ 0.70. (B) Mean cumulative incidence of AF and heart failure (HF)-related hospitalizations in subjects with AF stratified by presence of TTNmv. Hazard ratio (HR), 95% confidence interval (CI), and P-value were obtained from univariable Cox proportional hazard modeling. (C) Diagram denoting the location of TTNmv-T23756I. (D) Sequence alignment shows that the region of the T23756I is highly conserved across vertebrate species. (E) Allele frequencies of TTN-T3265I in various ethnic groups (gnomAD).

TTNmvs are associated with higher hospitalization risk

A total of 174 hospitalizations (64 AF-related and 110 HF-related) occurred after a median (interquartile range [IQR]) follow-up time of 4.14 (1.25-6.04) years, with 119 (68.4%) events occurring in TTNmv carriers and 55 (31.6%) of events occurring in non-TTNmv carriers. Thirty-nine (50.6%) subjects with a TTNmv experienced at least one hospitalization during the follow-up period compared to 22 (40.1%) of non-carriers, and total hospitalization incidence rate was 32.2 events per 100 person-years (p-y) in TTNmv carriers compared to 17.4 per 100 p-y in non-carriers. Mean cumulative incidence is shown in Figure 1B. The unadjusted hazard of hospitalization by 10 years was significantly higher in TTNmv carriers compared to noncarriers (hazard ratio [HR] 1.81, 95% confidence interval [CI] 1.04-3.15, p=0.036). This remained significant after partial adjustment for age and sex (HR 1.82, 95% CI 1.04-3.17, p=0.035) as well as full adjustment that additionally included ethnicity and baseline LVEF <50% (HR 1.80, 95% CI 1.03-3.15, p=0.039; Supplementary Table 4). No other covariates were significantly associated with the outcome. To assess whether baseline LVEF influenced the relationship between TTNmvs and hospitalization risk, an interaction term between LVEF<50% and presence of TTNmvs was tested and was not significant (interaction P=0.843). To assess whether prediction of deleterious effect influenced the relationship between TTNmv presence and hospitalization risk, subgroups based on high or low REVEL score were separately examined (Supplementary Table 5). Compared to non-TTNmv carriers, hospitalization risk was significantly increased in subjects carrying predicted deleterious TTNmvs (HR 1.92, 1.04-3.53, P=0.036). While subjects with TTNmvs not predicted to be deleterious also had an elevated point estimate of hospitalization risk, this difference was not significant (HR 1.60, 95% CI 0.78-3.28, P=0.198). Findings remained similar with exclusion of patients with nonischemic DCM (Supplementary Table 6). While in a previous study we identified likely pathogenic or pathogenic AF variants in 7.0% of ethnic minority probands with most (46.7%) TTNtvs11, these results suggest that TTNmvs are also linked with adverse clinical outcomes and missense variants should be considered in assessing pathogenicity in clinical and genetic contexts.

A single amino acid change in a TTNmv is potentially causative of AF

We identified three early-onset paroxysmal AF probands from separate families harboring the same rare TTNmv (Chr 2q31.2 (GRCh38): g. 178539797G>A). The clinical characteristics of the three probands are summarized in Table 2. This missense variant localizes in the A-band of TTN and causes the substitution of a conserved residue threonine by an isoleucine (p.Thr32756Ile, NM_001267550.2 (TTN), c.98267C>T (Figure 1C-D, Supplementary Figure 1A). In the gnomAD reference population, the TTN-T32756I variant has a frequency of 0.000245 overall, at a frequency of 0.0046 within the subpopulation of African ancestry and is impartially distributed across sexes and different ages (Figure 1E, Supplementary Figure 1B-E). The TTNmv has also been described in Clinvar (NM_001267550.2(TTN):c.98267C>T (p.Thr32756Ile), Supplementary Table 7) and gnomAD v4.0.0 (SNV:2-178539798-G-A(GRCh38)) in patients with DCM and other diseases.

TTN-T32756I iPSC-aCMs display aberrant contractility without affecting the sarcomere integrity

To functionally characterize the TTN-T32756I variant, we exploited wild-type (WT) human iPSC lines and introduced the point mutation c.98267C>T in the TTN allele (Figure 2A, Supplementary Figure 2A-B). The iPSC lines expressing the pluripotent markers Sox2 and Oct4 and showing normal karyotype were first differentiated into cardiomyocytes (Supplementary Figure 2C-D), and then a retinoic acid-based and comprehensive maturation protocol was applied to generate the iPSC-aCMs as we described previously26,27. Given that cardiac contractility is a key component of heart function and is linked to cardiac disorders associated with TTN38, we first assessed the contractility of the WT and mutant iPSC-aCMs and observed both reduced contraction and abnormal relaxation in TTN-T32756I-iPSC-aCMs (Figure 2B-E, Supplementary Figure 3A-C). Compared to WT, the beating frequency of the TTN-T32756I-iPSC-aCMs was significantly increased (52 ± 7.8 vs. 98 ± 7.5 beats per min, P=0.001; Figure 2C) coupled with the reduction of the contraction duration (456.5 ± 61.45 vs 262.9 ± 48.16 msec, P=0.032; Figure 2D), the peak-to-peak time (1529 ± 195.5 vs 636.6 ± 135.8 msec, P=0.004; Supplementary Figure 3B), and the relaxation (281.5 ± 42.95 vs 79.40 ± 21.14 msec, P=0.003; Supplementary Figure 3A). The contraction amplitude of the mutant was also increased (8.1 ± 0.8 vs 10.6 ± 7 au, P=0.040) without any significant changes in time-to-peak (Figure 2E, Supplementary Figure 3C), suggesting an increased contractile force by the TTN-T32756I-iPSC-aCMs. As sarcomere disorganization is often found in TTN-related cardiomyopathies and underlies contractile dysfunction39, we explored whether TTNmvs perturbed the organization of the sarcomere. Surprisingly, we did not find any disarray in the sarcomeres of either WT or TTN-32756I-iPSC-aCMs. Using transmission electron microscopy (TEM) and immunofluorescence (IF), we observed symmetrical arrays of recurring sarcomeres in the WT and the mutant atrial cardiomyocytes with no changes in the sarcomere length (Figure 2F-G, Supplementary Figure 3D-F). Although no sarcomeric perturbation by the T32756I variant was observed, the aberrant contractility displayed by the TTN-T32756I-iPSC-aCMs suggests that the TTNmv may be pathogenic.

Human-induced pluripotent stem cell-derived atrial cardiomyocytes (iPSC-aCMs) carrying TTN-T32756I variants exhibit atypical contractility but no sarcomere disorganization.

(A) Workflow for generating the CRISPR/Cas9-mediated iPSC line carrying the TTN-T32756I missense variation. (B-E) Contraction profile of wild type (black) and TTN-T32756I (Red) iPSC-aCMs (B), showing increased beating frequency (C), decreased contraction duration (D), and increased contraction amplitude (E) in the mutant. (F) Representative sarcomeric organization of wild-type (WT) and TTN-T32756I iPSC-aCM by Transmission electron microscopy (TEM). (G) There is no significant change in the sarcomere length. n.s. P>0.05; *P<0.05; **P< 0.01.

Altered cardiac potassium current and calcium handling underlies the arrhythmia phenotype in TTN-T32756I-iPSC-aCMs

As studies have shown that abnormal atrial electrophysiology and calcium-handling underlie atrial arrhythmogenesis and contractile dysfunction40,41, we hypothesized that the TTN-T32756I variant modulates atrial APs, ion channels and intracellular calcium. We first tested the spontaneous and paced AP characteristics of the isolated WT and the TTN-T32756I iPSC-aCMs, which revealed a significant shortening of the AP duration (APD) at 10%, 50%, and 90% repolarization, however, there was no change in the peak amplitude (Figure 3A-C, Supplementary Figure 4A-B). Since increased potassium current (IK), especially the augmented delayed rectifier potassium current (IKs), causes AF by reducing the APD27,42, we then assessed the IKs by whole-cell voltage clamping. We observed an increase in both IK and IKs in TTN-T32756I-iPSC-aCMs when compared to the WT (Figure 3D-F, Supplementary Figure 4C-E, Supplementary Figure 5A-D); thus, confirming that the APD shortening is partly due to the increased IK. To test the effect of T32756I on atrial calcium handling, we measured the intracellular calcium transients at the excitation-contraction coupling moment of the iPSC-aCMs using fluorescent (fura-2, AM) calcium imaging. Compared to the WT, spontaneously beating a TTN-T32756I-iPSC-aCMs exhibited increased frequency along with a significant reduction of the time to 50% and 90% decline of calcium transients (Figure 3G-I, Supplementary Figure 4F). We also saw a similar decline of calcium transients in the paced iPSC-aCMs (Supplementary Figure 5E-H). However, TTN-T32756I-iPSC-CMs exhibited similar calcium transient amplitudes as the WT, indicating that there was no change in the availability of the intracellular calcium for each contraction (Supplementary Figure 4G, Supplementary Figure 5). Our findings suggest that the reduced APD resulting from increased potassium current and the altered timing of the calcium transient may create a reentrant substrate for AF.

Effect of T32756I on action potential (AP) and calcium-handling in iPSC-aCMs.

(A-C) Representative optical AP recordings of spontaneously beating WT (A) and TTN-T32756I showing reduction of AP duration (APD) at the 50% (APD50) (B) and 90% (APD90) repolarization (C). (D) Current-voltage (I-V) curves of the slow delayed rectifier potassium current (IKs) in WT and TTN_T32756I iPSC-aCMs. control (n=7) (E-F) Comparison of IKs current density at 50 mV (mean ± SEM). (G) Representative tracings of spontaneous calcium transients of WT and TTN-T32756I iPSC-aCMs. (H-I) Calcium kinetics show that the TTN-T32756I iPSC-aCMs have increased frequency (H) and decreased transient durations (I) compared with the WT iPSC-aCMs. n.s. P>0.05; *P<0.05.

RNA sequencing analysis reveals alterations in cardiac signaling and disease pathways in TTN-T32756I-iPSC-aCMs

To elucidate the underlying molecular mechanism by which TTN-T32756I causes contractile defects and ion channel remodeling, we performed transcriptome sequencing of the mutant iPSC-aCMs and WT. Comparison of total RNA levels in TTN-T32756-iPSC-aCMs with WT-iPSC-aCMs and differential expression analysis showed genes related to cardiac muscle contraction and calcium handling were predominantly affected (Figure 4A, Supplementary Figure 6A-C). Gene-ontology (GO) pathway enrichment further showed that the top significantly downregulated cardiac-related GO Biological Processes (-BP) in TTN-T32756I iPSC-aCMs included key processes such as cardiac myofibril assembly, skeletal muscle myosin thick filament assembly, extracellular matrix organization, and regulation of potassium ion transmembrane transporter assembly among others (Figure 4B). As for GO-Molecular function pathways (GO-MF), outward potassium channel activity, voltage-gated potassium channel activity, calcium channel activity, and gap junction channel activity were enriched (Figure 4C). Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway analysis identified downregulation in critical pathways including Adrenergic signaling in cardiomyocytes, the Phosphoinositide 3-kinase (PI3K) signaling pathway, Dilated cardiomyopathy pathway, Hypertrophic cardiomyopathy pathway and Calcium signaling pathway (Figure 4D), which are crucial for normal cardiac electrophysiology and contractility. Upregulated pathways in both GO and KEGG analyses were mostly restricted to neuronal pathways including axon and neuron development, Hippo signaling pathway, and Notch signaling pathway. In the transcriptomic analysis of TTN-T32756I-iPSC-aCMs, several transcription factors (TFs) showed significant alterations, highlighting their potential roles in the mutation’s impact (Figure 4E, Supplementary Figure 6D). Top-cardiac related upregulated TFs include MYC, linked to cell growth and apoptosis, suggesting an increase in cellular proliferation. Stress-related TFs such as NFE2L2 and CEBPB are also elevated, indicating a heightened stress response. Additionally, TFs like ESRRA and FOXM1, which regulate metabolic processes and cell cycle respectively, along with FOXO1 for metabolism and stress resistance, were upregulated. Key TFs in muscle function such as MYOD1, MYOCD, KLF5, and NKX2.5 were enhanced, alongside SMAD2, CREBBP, PITX2, NFKB1, and GATA4, which are vital for diverse roles ranging from signal transduction to immune regulation (Figure 4E). Conversely, downregulated transcription factors include SIRT1, which plays a role in cellular stress resistance and mitochondrial function, pointing to potential vulnerabilities in cellular defenses. Cardiovascular-related TFs such as KLF2 and KLF3, along with SMAD7, involved in TGF-beta signaling, were reduced, possibly impacting structural and signaling integrity. Additionally, MXD1 and IKZF1, crucial for cellular differentiation and immune development, were also diminished (Figure 4E). These transcriptional changes underscore a complex network of regulatory adjustments that could contribute to the altered electrophysiological and structural properties in TTN-T32756I-iPSC-aCMs.

Transcriptomic profile and pathway enrichment analysis comparing TTN-T32756I iPSC-aCMs with the WT.

(A) Heatmaps of cardiac-related upregulated and downregulated differentially expressed genes (DEGs). (B) Top significantly enriched downregulated cardiac-related Gene-Ontology Biological process (GO-BP) pathways in the TTN-T32756I iPSC-aCMs. (C) Top significantly enriched downregulated cardiac-related Gene-Ontology Molecular Function (GO-MF) pathways in the TTN-T32756I iPSC-aCMs. (D) Top significantly enriched downregulated cardiac-related Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways in the TTN-T32756I iPSC-aCMs. (E) Significantly enriched upregulated and downregulated transcription factors (TFs). (F) Network diagram showing the upregulation of KCNQ1 by FHL2 predicted by the Ingenuity pathway enrichment analysis (IPA).

Utilizing the KEGG and GO pathways, we performed Ingenuity Pathway Analysis (IPA) to identify novel titin-interacting proteins that could cause the increased IKs activity in the TTN-T32756I-iPSC-aCMs. The IPA suggests FHL2, which is known to transduce mechanical signaling through its interactions with titin and is enriched in the mutant iPSC-aCMs (Figure 4F). As FHL2 has also been shown to directly interact with the IKs-binding partner mink (KCNE1) and increases its current density43, we postulate that elevated FHL2 level in TTN-T32756I-iPSC-aCMs may cause enhanced binding to IKs, and thus increase its activity.

The enhanced IKs current in TTN-T32756I-iPSC-aCMs is mediated by the FHL2

To determine if TTN-T32756I increases IKs by modulating the interaction between KCNQ1-KCNE1 and FHL2, we performed co-immunoprecipitation studies and confocal microscopy in both WT and TTN-T32756I-iPSC-aCMs. The co-localization between KCNE1 and FHL2 increased ∼3-fold in TTN-T32756I-iPSC-aCMs, suggesting an increased interaction between them (Figure 5A, Supplementary Figure 7). Since FHL2 enhances IKs activity, we then investigated whether inhibition of FHL2 could reverse the increased IKs that was observed in the TTN-T32756I-iPSC-aCMs. To inhibit FHL2, we employed the small interfering (si) RNA to suppress the FHL2 gene and measured the IKs activity in both WT and TTN-T32756I-iPSC-aCMs. Both the WT and mutant iPSC-aCMs transfected with FHL2-specific siRNA showed a substantial decrease in FHL2 expression (Figure 5B). As shown in Figure 5C-D, voltage-clamp recordings demonstrate that the IKs in FHL2-suppressed TTN-T32756I-iPSC-aCMs are significantly reduced compared to the corresponding IKs in TTN-T32756I-iPSC-aCMs but comparable to the WT (Figure 5D). Therefore, inhibition of FHL2 by the siRNA rescues the increased IKs of the TTN-T32756I-iPSC-aCMs, which is similar to that of WT IKs. Overall, our data suggest that the TTNmv creates an EP substrate for AF by modulating the IKs activity in part by an increased interaction between the KCNQ1-KCNE1 complex and FHL2 (Figure 5E).

Inhibition of FHL2 rescues enhanced IKs in TTN-T32756I iPSC-aCMs.

(A) Co-immunoprecipitation revealed increased interaction between FHL2 and KCNQ1-KCNE1 (MinK) complex. (IP: KCNE1). Immunoblotting (IB) was performed with antibodies against FHL2 (32 kDa) and MinK (32 kDa) (n = 3) (B) qPCR data showing the inhibition of FHL2 gene by the siRNA in the WT and TTN-T32756I iPSC-aCMs (n=7). (C) I-V curves showing the rescue of the IKs TTN-T32756I iPSC-aCMs by the suppression of FHL2 (n=4-8). (D) Comparison of IKs current density at 50 mV (mean ± SEM). (E) Schematic showing the TTN-T32756I results in increased FHL2 binding with the KNCQ1-KCNE1 complex and enhanced IKs activity. n.s. P>0.05; *P<0.05; ***P<0.001.

Discussion

We identified an association between TTNmvs and clinical outcomes among a multiethnic cohort of AF patients and elucidated a causal mechanism by which a TTNmv may lead to AF. Although TTNtvs are a well-recognized cause of DCM and have been associated with early-onset AF10,44, the role of TTNmvs in either arrhythmia or cardiomyopathy pathogenesis is unknown. Here, for the first time, we show that TTNmvs are associated with increased hospitalization risk in a multiethnic AF cohort, and that a TTNmv can cause AF by impairing contractility and ion channel remodeling in atrial cardiomyocytes. Furthermore, we demonstrate that augmented IKs mediated by FHL2 can create a substrate for AF. Together, these findings establish a potential causal role of TTNmvs in AF and support IKs and FHL2 as potential therapeutic targets for TTN-related AF.

The genetic architecture of AF is complex, with both monogenic and polygenic contributions that intersect with age and comorbidities to define an individual’s risk. Despite AF being traditionally considered an ion channelopathy, the sarcomeric gene, TTN, is the most common in which rare loss-of-function variants have been associated with AF10,45, and penetrance of such variants was found to be even greater for AF than for heart failure in the UK Biobank46. As rare TTN variants comprise the strongest monogenic contribution to AF risk, we sought to explore whether TTNmvs may similarly have implications in AF development or outcomes. A study of 147 probands with DCM identified 44 severe TTNmvs in 37 probands, which clustered in the A-band region17. There were no differences in heart transplant-free survival between carriers and noncarriers. In a more recent study of 530 subjects with DCM using more stringent allele frequency criteria, 31 predicted deleterious variants were identified, also predominantly located in the A-band19. However, these TTNmvs were found with similar frequency to reference populations. Finally, a study of two families with DCM demonstrated segregation of TTNmvs affecting the same highly conserved cysteine residue in the I-band, which showed impaired contraction and folding at physiological temperatures in a homozygous iPSC model, demonstrating that a TTNmv can cause DCM23.

We identified TTNmvs in 58% of subjects in our cohort, and 33% carried a potentially deleterious variant by REVEL in silico prediction. This frequency was higher than the 37/147 (25.2%) of TTNmv carriers in a DCM cohort17, understanding that differing methods of bioinformatic filtering may limit ability for direct comparison. Aligning with prior studies, most deleterious variants in our study were found in the A-band47,48, where TTNtvs have been linked most strongly to disease. Presence of predicted deleterious TTNmvs was associated with higher LV end diastolic diameter but without statistically significant differences in LVEF, potentially suggesting overlap between a subclinical left ventricular cardiomyopathy and AF in these subjects. We additionally noted higher ventricular rate and QTc interval on index ECGs in these subjects. Rare deleterious variants in TTN have previously been associated with changes in the QT interval49, but the mechanism for this remains unclear. Finally, evidence on the relationship between TTNmvs and clinical outcomes is sparse, with limited studies examining event-free survival in DCM patients. We observed a significant association between TTNmv presence and increased cumulative incidence of AF or HF-related hospitalizations, which remained unchanged in multivariable adjustment and in sensitivity analyses. This importantly suggests that TTNmvs may correlate with disease severity. Our cohort is derived from a single-center multi-ethnic registry of individuals with AF and lacks a matched cohort of non-AF controls to compare the incidence of TTN missense variants. Further study exploring these associations in mult-ethnic, larger validation cohorts that include both AF and non-AF individuals and examining AF-specific measures such as arrhythmia burden or treatment response will be necessary to fully understand the clinical importance of TTNmvs in AF.

Prediction of missense variant impact using in silico tools and algorithms can be limited by low specificity, contributing to challenges in clinical interpretation50. Functional studies are often helpful in these cases to confirm or support variant pathogenicity. As existing in vitro or in vivo models do not adequately replicate the complexity of AF, human iPSC-aCMs possess the complex array of cardiac ion channels that make up the atrial AP and provide an adequate model to establish causal relationships in AF26. We identified the missense variant TTN-T32756I in three unrelated subjects in our multiethnic AF registry. Hypertension is a common comorbidity in patients with AF and could contribute to disease progression. However, all three individuals carrying TTN-T32756I exhibited early-onset AF (onset before 66 years), with one case occurring as early as 36 years. This suggests a potential two-hit mechanism, where genetic predisposition and comorbidities influence disease risk. Importantly, our iPSC model isolates the genetic effects of TTN-T32756I from other factors, supporting a direct pathogenic role. Remarkably, the single amino acid change in this giant protein titin results in aberrant contractility and EP remodeling without affecting the sarcomeric integrity. Although we did not observe any defects in the sarcomere, in a separate study we showed that deletion of nine amino acids including the T32756I in the same area caused significant perturbation in the sarcomere assembly37. T32756 corresponds to a conserved Threonine of the Ig139 domain (UniProt# Q8WZ42, predicted Ig155 by TITINdb2) and the replacement of a conserved hydrophilic residue (Thr) in WT with a hydrophobic residue (Ile) may cause thermal instability and destabilize the domain. It is now well recognized that point mutations may drastically disrupt the Ig domains and could possibly unfold under pathological conditions51. Such mutations have been proposed to modify titin-based passive stiffness and may lead to cardiomyopathy. Missense mutations such as T2850I, R57C, S22P in the I-band titin have been reported to significantly destabilize Ig domains and display a higher tendency for unfolding52,53. Likewise, T32756I may unfold under the physiological settings and increase susceptibility to protein loss (haploinsufficiency) or degradation (poison peptide effect), leading to impaired sarcomere function. Furthermore, as titin acts as a crucial signaling hub that transduces mechanical forces to downstream signaling pathways within cardiomyocytes and interacts with various sarcomeric proteins and signaling molecules13,54, subtle destabilization of Ig139 domain by the T32756I could alter the biochemical signals that regulate various downstream signaling pathways involved in cardiac function and adaptation. Hence, the contractile and ion channel dysfunction caused solely by the homozygosity of T32756I in the iPSC-aCMs provide strong evidence that the AF is exclusively caused by the TTN-T32756I missense variant.

Although several studies have identified possible pathways associated with TTNtvs and AF, the pathogenic mechanisms of TTNmv-associated AF remain unclear. We show for the first time that the T32756I creates an EP substrate for AF by ion channel remodeling with an enhanced IKs, which is mediated by a titin interacting protein FHL2. FHL2 is enriched in cardiac muscle and is a multifunctional protein that regulates cardiac myocyte signaling and function55. FHL2 interacts with KCNE1 that binds to the outer face of the KCNQ1 channel pore domain and may modify the interactions between the voltage sensor, S4-S5 linker, and the pore domain to augment the channel current43. It has been suggested that FHL2 interacts with titin N2B segment, titin kinase domain and forms a complex with other proteins, such as MURF1 and MURF2, calmodulin, and Nbr1, playing a role in hypertrophy signaling and the atrophy response56,57. It also binds to the edge of titin (M-line), which is in close proximity to T32756I, and interacts with myospryn and obscurin58. It is tempting to postulate that T32756I may interfere with FHL2’s ability to interact with titin, which could then disrupt with subsequent mechano-biochemical signaling and increase FHL2’s availability for binding with the KCNQ1-KCNE1 complex. While our study provides mechanistic insights into the role of the TTN-32756I missense variant in AF, TTNmvs may be involved in the regulation of ion channel modulation and atrial rhythmicity through a multitude of interrelated signaling pathways. Further studies are needed to fully uncover the molecular basis of this regulation and its implications for AF pathophysiology and potential therapeutic interventions.

In summary, this study supports a causal relationship between a TTNmv and AF and suggests that TTNmvs may be associated with adverse outcomes in patients with AF. The study also unveils the critical role of FHL2 in modulating the atrial action potential via binding to Kv7.1-KCNE1 and enhancing the IKs in TTN-T32756I-iPSC-aCMs. Importantly, we found that the increased IKs can be rescued by FHL2 suppression. This suggests that targeting IKs or FHL2 may restore and maintain sinus rhythm and improve atrial contractility. Further study is needed to validate the proposed mechanism and determine if TTNmvs in other regions are associated with AF by a similar process. Our data indicate that AF-associated TTNmv can cause arrhythmia, provide new insight into the mechanisms underlying AF and suggest that targeting KCNQ1 or FHL2 could represent a new therapeutic strategy for improving cardiac function. While we have not yet explored the effect of TTN-T32756I in iPSC-derived ventricular cardiomyocytes, it would be interesting to investigate whether this variant produces similar or distinct EP alterations in the ventricular cardiomyocytes. Our findings may also have broader implications for the treatment of patients harboring disease-causing rare variants in sarcomeric proteins and suggest that genomic analyses that encompass TTNmvs should also consider experimental evidence in advancing our understanding of AF etiology and improving patient care in the era of precision medicine.

Methods

Study population

The UIC Multi-Ethnic AF Biorepository was established in 2015 to explore the genetic basis of AF across race and ethnicity. This study was approved by the UIC Institutional Review Board (#2015-0681). Subjects are prospectively enrolled from outpatient and inpatient sites within the University of Illinois Health (UIH) healthcare system. Subjects must have a documented history of AF by electrocardiogram, Holter/event monitor, or implantable cardiac device, with confirmation by an attending cardiologist. At time of enrollment, blood is drawn for DNA extraction and genetic analysis. Baseline demographics are obtained through provider notes in the electronic health record (EHR) and confirmed with the patient at the time of enrollment as necessary. Longitudinal clinical outcomes are followed through serial EHR review. Written informed consent was obtained from all participants. Adults 18 years of age or older at time of AF diagnosis were prospectively enrolled between August 25, 2015, and May 19, 2019. Samples from 161 subjects who identified as non-Hispanic Black (NHB) or Hispanic/Latinx (HL) and had an echocardiogram performed within 3 months of enrollment date underwent whole exome sequencing. A total of 27 subjects with congenital or rheumatic disease, severe mitral stenosis, and end stage renal disease on dialysis were excluded. Three subjects who carried a predicted loss-of-function frameshift or stop-gain TTN variant were additionally excluded. The primary exposure was the presence of one or more missense TTN variants. The primary outcome was time from initial AF diagnosis to hospitalizations with a primary diagnosis of either AF or acute decompensated HF recorded in the clinical discharge summary, captured as repeating events. Clinical data was obtained through manual review of EHR and comorbidities were defined if present in clinical notes on or prior to AF diagnosis date. Subjects were censored on date of death, last known clinical encounter (up to August 26, 2023), or at 10 years after AF diagnosis. Echocardiographic criteria for left ventricular dilatation was defined as left ventricular end diastolic diameter (LVEDD) > 2 standard deviations (SD) above sex-specific mean according to standard guidelines59. Subjects meeting echocardiographic criteria for left ventricular dilatation fulfilled clinical criteria for nonischemic dilated cardiomyopathy if left ventricular ejection fraction (LVEF) < 50% and ischemic cardiomyopathy was ruled out by coronary angiogram.

Whole exome sequencing

Samples were sequenced with support of the National Human Genome Research Institute (NHGRI) Centers for Common Disease Genetics (CCDG) program. Sequencing was performed at the Broad Institute following methods established by the National Heart, Lung, and Blood Institute (NHLBI) Trans-Omics for Precision Medicine (TOPMed) Atrial Fibrillation Study (available under TOPMed Whole Genome Sequencing Methods: Freeze 9 at https://www.ncbi.nlm.nih.gov/projects/gap/cgi-bin/study.cgi?study_id=phs001062.v5.p2). Samples underwent WES using an Illumina HiSeq X system with 150bp paired-end read length, at 20x depth for at least 85% of targets Alignment to human genome reference GRCh38/hg38 was performed using BWA-MEM (Burrows-Wheeler Aligner, v0.7.15.r1140). TTN variant calls were filtered to select those with read depth of ≥20X, genotype quality ≥20, minor allele frequency (MAF) ≤1% across gnomAD subpopulations, and excluding intronic, regulatory or UTR 3’/5’ variants. Synonymous variants, homozygous reference calls, non-canonical variants, and any variants with percent spliced in (PSI) index <90 were excluded. Missense variants predicted to be deleterious were identified using REVEL score of ≥0.7, a previously proposed cutoff to predict deleterious effect in dilated cardiomyopathy60,61. Because of the uncertain role of missense TTN variants across AF and other cardiac diseases, we examined all TTN missense variants regardless of REVEL score. Variant annotation data were obtained from Ensembl Variant Effect Predictor62, downloaded January 11, 2024, and CardioDB (https://cardiodb.org/titin) for TTN region and percent spliced in (PSI)63, downloaded November 16, 2023. The TTN-T32756I variant (REVEL Score: 0.58758, Supplementary Table 1) was prioritized due to its occurrence in multiple unrelated individuals within our clinical AF cohort, despite no reported family history of AF in affected individuals. While no parental inheritance was observed, the possibility of de novo origin cannot be excluded. Furthermore, this variant is located within a region overlapping a deletion mutation recently shown to cause AF in a zebrafish model supporting its potential pathogenicity37. Notably, the affected individuals did not carry additional loss-of-function TTN variants.

Human iPSC culture and human iPSC-aCM differentiation

Human iPSC-CMs were derived from reprogrammed peripheral blood mononuclear cells (PBMCs) as previously described29. 80-90% confluent iPSC-CMs was differentiated using the Cardiomyocyte Differentiation Kit (Gibco) and guided toward the atrial subtype using all-trans RA64. The cellular population was purified through glucose starvation and lactate replacement, resulting in contracting monolayers of iPSC-aCMs. Our protocol typically yields ∼80 to 90% pure iPSC-aCMs and <6% fibroblasts based on immunostaining analysis as we have previously described29,64. Human iPSC-aCMs were then matured following dissociation and replating on fibronectin-coated plates and maintained in Cardiomyocyte Maintenance Media supplemented with T3, insulin-like growth factor-1, and dexamethasone as previously described27.

Generation of TTN-T32756I hiPSCs using CRISPR-Cas9

Due to the patients’ unavailability of peripheral blood mononuclear cells (PBMCs), we utilized a healthy iPSC line and introduced the TTN-T32756I variant using CRISPR/Cas9 genome editing. This approach ensures an isogenic background, thereby minimizing genetic variability and providing a controlled system to study the direct effects of the mutation. We edited the genome of WT iPSCs to introduce the T32756I using CRISPR-Cas9 technique29. The TTN-T32756I CRISPR-Cas9 was designed from WT allele sgRNA and two ssODN that target the TTN genomic locus: 31111-31119 and genomic DNA location: chr2:178539785-178539811 (GRCh38.p13). Amplification in the exon 352 cDNA donor RNP was constructed and electroporated into the TTN gene WT hiPSCs. Gene-editing efficiency was confirmed with next-generation sequencing (NGS). Karyotype analysis was carried out by the Cytogenetics laboratory at WiCell Research Institute Inc. Cells were collected and chromosome were evaluated using the giemsa trypsin wright (GTW) banding method. Metaphase cells were analyzed, all of which were concluded to have a normal karyotype (46, XY)

Contraction analysis and optical voltage mapping

Contractility of hiPSC-aCMs was performed with a versatile open-source software, MUSCLEMOTION65. MUSCLEMOTION is ab ImageJ plugin that utilizes a video-based system to assess contractile functions. Optical voltage mapping recordings were performed on the IonOptix system myopacer system using the fluovolt membrane potential kit (Thermo Fisher). HiPSC-aCMs were cultured in confocal dishes were incubated with Tyrode’s solution (140 mM NaCl, 4.56 mM KCl, 0.73 mM MgCl2, 10 mM HEPES, 5.0 mM dextrose, 1.25 mM CaCl2) plus 1x Fluovolt (Sigma/Aldrich) for 15-20 minutes. Cells were then rinsed with normal Tyrode’s solution and recorded APDs on the IonOptix system. For pacing experiment, iPSC-aCMs were electrically stimulated at a frequency of 0.5 Hz with platinum wires separated 2-4 mm, stimulation voltage was 25 % above threshold (10 V, 4 msec). These experiments were performed at the room temperature (25°C).

Electrophysiology

Whole-cell patch clamping on hiPSC-aCMs IK recordings was performed according to previously published protocols27,29. Briefly, voltage-clamps were achieved by using an Axopatch 200B amplifier controlled by pClamp10 software through an Axon Digidata 1440A. External solution for IK recordings contained: 140 mM NaCl, 4 mM KCl, 1.8 mM CaCl2, 1.2 mM MgCl2, 10 mM glucose, 10 mM HEPES, and 0.01 mM nifedipine, adjusted to pH 7.4 with NaOH. IKs recordings were isolated as 1 μM HMR-1556–sensitive current. The intracellular solution contained 100 mM potassium aspartate, 2 mM MgCl2, 20 mM KCl, 5 mM Mg-ATP, 5 mM EGTA, and 10 mM HEPES adjusted to 7.2 with KOH. IKs currents were elicited by using 3-second voltage-clamp steps to test potentials of −60 to +60 mV from holding potential of –40 mV and with 20 mV increments.

Calcium Handling

Calcium transients were measured using Fluo-4-AM (Invitrogen) dye dissolved in 2.5% Pluronic F-127 (MilliporeSigma). To get a working concentration of 5 μM, the dye solution was added to Tyrode’s solution (1 mM Ca2+). The cells were treated with Fluo-4-AM in 1 mM Ca2+ Tyrode’s solution, and then allowed to sit at room temperature for 20 minutes in the dark. Tyrode’s solution without indicators and containing 2 mM calcium was used to wash the cells. Using a 40× objective and the Zeiss LSM 710 confocal equipped with a BiG module, line scans were acquired and examined using ImageJ. The corrected minimum and maximum fluorescence values were determined by normalizing the fluorescence using a baseline background region that was unique to each cell. Fluorescence signals were normalized to basal cell fluorescence after fluo-4 loading (F0). For pacing experiment, iPSC-aCMs were incubated at room temperature with 1 μM Fura 2-AM for 10 min and then washed out with normal tyrode 1.8 mM Ca for 20 min. The cells were then electrically stimulated at a frequency of 0.5 Hz with platinum wires separated 2-4 mm, stimulation voltage was 25 % above threshold (10 V, 4 msec). Estimations of intracellular Ca2+ are reported as changes in ΔF/F0, where ΔF = F-F0. Calcium transients’ amplitude (340/380 nm ratio, and kinetics were analyzed using IonOptix software.

Transmission electron microscopy (TEM)

iPSC-aCMs were fixed with 2.5% glutaraldehyde in 0.1M Sorenson’s Buffer for 60 minutes at room temperature. The cells are then carefully collected into a microcentrifuge tube pre-filled with same the fixation buffer. The samples were then centrifuged at 2500xg for 10 min at room temperature. The pellet was then removed and inverted with a hypodermic needle to ensure fixative solution thoroughly permeates the sample. The pellet is then allowed to incubate at room temperature to fix for an additional 60 minutes at room temperature. The fixation solution was then removed and substituted with 1% glutaraldehyde + 4% paraformaldehyde in 0.1M Sorenson’s buffer and stored in 4°C. Fixed samples were embedded in resin and microtome sections were imaged on a JEOL JEM-1400 Flash TEM.

Immunofluorescence

Cells were cultured on Matek(R) glass bottom dishes that were coated with vitronectin. The cells were then allowed to grow for a period of 2 days. The cells were then fixed with 4% paraformaldehyde (PFA) for 10 minutes and permeabilized with phosphate-buffered saline (PBS) containing 0.1% Triton X-100 for 10 minutes. After that, the cells were blocked with 3% bovine serum albumin (BSA) in PBS for 1 hour at room temperature. The staining procedure was carried out overnight at 4°C using primary antibodies that were diluted in 3% BSA in PBS. The cells were rinsed three times with PBS and then exposed to secondary antibodies (anti-mouse-FITC or anti-rabbit-PE, Santacruz, 1:1000) for one hour at room temperature. Subsequently, the cells were rinsed 3x with PBS and co-stained with 1µg/ml of DAPI. The antibodies Anti-SOX2 (Abcam, 1:200) and anti-OCT4 (Santacruz, 1:200) were used to assess the generation of iPSCs, while anti-cTnT (Invitrogen, 1:300) and anti-Kv1.5 (Alomone labs, 1:200) were utilized to confirm the differentiation of atrial cardiomyocytes. The samples were visualized using an LSM710 Meta Confocal Microscope manufactured by Zeiss.

Protein isolation, western blotting, and co-immunoprecipitation

We performed Western blots as previously described27,29. For western blots, cells on 6-well plates were washed with ice cold DPBS without Ca2+ and Mg2+, after which 250 µL of 1X RIPA with protease and phosphatase inhibitors was added per well. Lysate concentrations were measured using BCA Assay, and diluted with 4X Laemmli buffer with 10% 2-Mercaptoethanol. Per sample, 25 µg of protein was then run on an SDS-polyacrylamide gel to separate, and resolved gels were electro-transferred on 0.2 µm PVDF membranes. Membranes were blocked with 5% BSA for 1 hour, and then probed with corresponding antibodies of target proteins (anit-FHL2 antibody, Abcam#ab202584). The blots were developed using either anti-rabbit HRP or anti-mouse HRP and scanned on C280 imaging systems (Azure Biosystems). Protein signal densities were determined using ImageJ and normalized to corresponding β-actin signal densities.

For co-immunoprecipitation experiments, cells are rinsed with ice cold DPBS without Ca2+ and Mg2+, after which 500 µL of 1X RIPA with protease and phosphatase inhibitors was added each well of a 6-well plate. Lysates are sonicated on ice 3×5 seconds, then centrifuged for 10 min at 14,000xg at 4°C. 100µL of Protein A/G agarose bead slurry was added to the lysate and incubated at 4°C on a rotator for 60 minutes. The cell lysate with bead slurry was centrifuged for 10 minutes at 1000xg at 4°C and the supernatant was transferred to a fresh Eppendorf tube. 2µg of primary antibody (KCNE1 [Alomone# APC-168]: 2.5µL) was added to 500µg of precleared cell lysate, and the mixture was incubated with gentle rotation overnight at 4°C for a day. 40µL of Protein A/G bead slurry was the washed 3x with RIPA, and the precleared cell lysate with the primary antibody was added to the bead slurry, and then followed by incubation for 3 hours at 4°C. The mixture is centrifuged for 10 min at 1000xg at 4°C, and the flow through was saved to verify the immunoprecipitation. The bead pellet was rinsed with 5×500 µL RIPA and resuspended in 30µL 4X Laemmli sample buffer, vortexed, then centrifuged for 30 seconds. The sample was then boiled for 5 min, then centrifuged at 14,000xg for 5 min, representing the IP sample. 15 µL of supernatant and immunoprecipitate samples were loaded onto a 10% SDS-PAGE gel and analyzed by western blotting.

RNA-Sequencing

The RNA was isolated using the Maxwell® RSC simplyRNA Cells Kit (Promega AS1390) according to the instructions provided by the manufacturer. Library preparation was carried out using the Universal Plus mRNA-Seq kit (NuGen 0520-A01). Briefly, RNA underwent poly-A selection, enzymatic fragmentation, and generation of double-stranded cDNA using a mixture of Oligo (dT) and random priming. The cDNA was subjected to end repair, ligation of dual-index adaptors, strand selection, and 15 cycles of PCR amplification. The concentrations of the purified library were determined using the Qubit 1X dsDNA HS Assay Kit (Invitrogen Q33231). The libraries were then combined in equal amounts, considering the Qubit concentration and the average size determined by TapeStation. The pooled libraries were subsequently run on the MiniSeq instrument for index balancing. The ultimate, refined pool was measured using quantitative polymerase chain reaction (qPCR). The raw reads were aligned to the reference genome hg38 using the STAR alignment tool. The quantification of ENSEMBL gene expression was performed using FeatureCounts. The edgeR package was used to calculate normalized and differential expression statistics. To account for multiple testing, p-values were adjusted using the false discovery rate (FDR) correction method. We detected differentially expressed genes (DEGs) with a q value of less than 0.05 and a log 2 fold change of at least 2 between the Control and AF iPSC-aCMs. We conducted unsupervised hierarchical clustering of all genes that showed differential expression using the Euclidean distance and complete linkage method. Additionally, we created volcano plots using the R programming language. The up-and down-regulated genes were analyzed individually using the DAVID functional annotation tool against the Gene Ontology Biological Process (GO BP) database.

qPCR

Total RNA was isolated from hiPSC-aCMs using TRIzol reagent (Invitrogen), following the manufacturer’s instructions to ensure the extraction of high-quality RNA. The concentration and purity of the isolated RNA were carefully assessed using a NanoDrop 2000 spectrophotometer (Thermo Fisher Scientific), with 1 µg of total RNA utilized for each reverse transcription reaction. Reverse transcription to synthesize cDNA was conducted using SuperScript III Reverse Transcriptase (Thermo Fisher Scientific). For the analysis, specific assays and primers were selected for target gene FHL2 (Forward sequence: GTGGTGTGCTTTGAGACCCTGT, Reverse sequence: GAGCAGTGGAAACAGGCTTCATG) with glyceraldehyde 3-phosphate dehydrogenase (GAPDH) serving as the normalization reference gene. qPCR reactions were performed on an ABI QuantStudio 5 system (Applied Biosystems), using SYBR Green PCR Master Mix to accurately detect and quantify PCR amplification products. Relative expression levels of the target genes were calculated employing the ΔΔCt method, by the quantification of gene expression changes in the experimental samples relative to control. For the ΔCT Calculation, the cycling time (CT) value of the target gene was subtracted from the CT value of GAPDH in the same sample using ΔCT=CTtarget gene − CTreference gene. The ΔΔCT value was then calculated using ΔΔCT= ΔCTExperimental – ΔCTControl. The relative expression for the gene was in turn calculated using Relative gene expression= 2−ΔΔCT.

siRNA experiments

FHL2-specific siRNAs (#SR301594) or scrambled siRNAs were applied to mature hiPSC-aCMs using Lipofectamine RNAiMAX (Invitrogen, Carlsbad, CA). Opti-MEM medium was used to dilute stock solutions of lipofectamine and siRNA, each prepared at a concentration of 10 µM. The siRNA-lipid complex was then made by mixing these solutions in a 1:1 ratio and incubated for five minutes. After adding this complex to the cells dropwise, the media was changed, and the cells were allowed to incubate for a further two days.

Data Analysis and Statistics

For clinical data, categorical variables are represented as count and percentage (%) and tested by Fisher’s exact test. Continuous variables are reported as mean (standard deviation [SD]) or median and interquartile range (IQR) where specified and tested with Mann-Whitney U and Kruskal-Wallis tests. Ordinal variables were tested with Kruskal-Wallis test. Univariable and multivariable Cox proportional hazards models were used to evaluate risk of the primary outcome as recurring events, related to presence of TTN missense variants. A partially adjusted model was first tested using covariates of age and sex, followed by a fully adjusted model which additionally incorporated covariates of ethnicity and baseline LVEF <50%. Interaction between TTN missense variants and LVEF was additionally evaluated. To identify whether associations were limited to variants with predicted deleterious effect, analysis was repeated stratifying subjects according to presence and predicted impact of TTNmv: presence of one or more predicted deleterious (REVEL≥0.7) variants, presence of predicted benign (REVEL <0.7) variants only, or no TTNmv present. Additional sensitivity analyses were performed excluding subjects with multiple TTNmvs and with nonischemic dilated cardiomyopathy. Analysis was performed in R (version 4.2.1, packages: arsenal, ggplot2, ggsurvfit, gtsummary, reda, survival).

Experiments were performed at least two times (biological replicates) to ensure reproducibility. Unless otherwise noted, experimental data on hiPSCs are shown as mean ± SD. *P < 0.05, **P < 0.01, ***P < 0.001, ****P < 0.0001 indicate significance, while P > 0.05 is regarded as non-significant. Statistical analyses include unpaired nonparametric and 2-tailed Mann-Whitney U test for data with normal distribution, and either 1-way or 2-way ANOVA with post hoc Bonferroni’s corrections for several groups. The first and third quartiles of a median are used to express skewed data. Fisher’s Exact test is used to compare categorical data, and unpaired Student’s t-test or ANOVA are applied to assess continuous variables.